\section{The local WSPD death-time implementation}\label{sec:wspd}

This appendix discharges the one implementation step deferred by \cref{sec:dt}: realizing a locally
navigable Euclidean $(1+\rho)$-spanner of a point set $X\subseteq[\Delta]^2$ under the range-counting
oracle, and running the clipped death-time sampler of \cref{sec:dt-sampler} on it. We build on the
WSPD construction of Driemel et al.~\cite{DMOSW25} (whose spanner
facts are from Har-Peled~\cite{HarPeled11} and whose sampling primitive is from
Monemizadeh~\cite{Monemizadeh23}). We use \emph{cell} for a node of the quadtree $\mathcal T$ on
$[\Delta]^2$; a cell $c$ has side length $r(c)$ and diameter $\sqrt2\,r(c)$, and $d(c,c')$ is the distance
between cell centers. The vertex set is $X$; for $X=P$ a cell is nonempty when $\rcount\cdot>0$. For a
snapped support $X=Q$ we take the grid side $h$ to be a dyadic quadtree scale, so each support $h$-cell is
itself a quadtree cell and every region queried below is a union of $h$-cells; an $h$-cell is then nonempty
exactly when the aligned $P$-region is nonempty, so every emptiness test is one count on $P$
(\cref{sec:dt}). Throughout, the WSPD separation is $1/\rho$ (a
pair is well separated when its cells are at distance $\ge\max\{r(c),r(c')\}/\rho$), chosen so that the
spanner stretch below is $1+\rho$.

\paragraph{The construction at a glance.}
The plan has four layers, each a lemma below. (1) A uniform vertex of $X$, and the representative of any
cell, are drawn with $O(\log\Delta)$ counts by telescoping sampling (\cref{lem:telescoping}). (2) The
spanner $S$ has one edge per pair of a local family $\widehat{\mathcal W}$, joining the two representatives at their exact Euclidean
distance; because we know representative coordinates, edge weights are free and $S$ is a genuine Euclidean
subgraph (\cref{sec:wspd-spanner}). (3) From a vertex $u$, its incident spanner edges below a length are
enumerated by scanning $u$'s $O(\log\Delta)$ quadtree ancestors and, for each, its $O(\rho^{-2})$
well-separated partner cells (\cref{lem:incident}). (4) The clipped death time of a seed is computed by an
exact bottleneck Dijkstra (\cref{alg:deathtime}) that halts at the first lower-ranked or root vertex after
$O(\log\abs V)$ expansions in expectation. A single global query budget, in place of a per-search cap,
controls the worst case without biasing any estimate. Nothing here reveals a point's location beyond what
a representative computation already costs, and every line is a range count.

\subsection{Uniform point sampling under the oracle}\label{sec:prelim-sample}

We first record the sampling primitive used here and in \cref{sec:support} (referenced from
\cref{sec:support-oracle}). It is \emph{telescoping sampling}~\cite{Monemizadeh23}.

\begin{lemma}[Uniform sampling]\label{lem:telescoping}
Given a single quadtree cell $c$ with $\rcount c>0$ (in particular the root $[\Delta]^2$), one can return a
point of $P\cap c$ drawn uniformly at random using $O(\log\Delta)$ range-counting queries.
\end{lemma}

\begin{proof}
Walk down $\mathcal T$ from $c$: at a cell with children $c_1,\dots,c_4$, query the four counts
$\rcount{c_i}$ (which sum to the current count) and descend into $c_i$ with probability proportional to
$\rcount{c_i}$. After $O(\log\Delta)$ levels the walk reaches a single grid point $p$. The probability of
reaching $p$ is the product of the conditional probabilities along its root-to-leaf path within $c$, which
telescopes:
\[
  \Pr[\text{reach }p]=\prod_{\text{levels}}\frac{\rcount{c^{(\ell+1)}}}{\rcount{c^{(\ell)}}}
   =\frac{\rcount{\{p\}}}{\rcount c}=\frac1{\abs{P\cap c}},
\]
where $c^{(\ell)}$ are the cells on $p$'s path from $c$; so $p$ is uniform on $P\cap c$. Each level spends
$O(1)$ counts and there are $O(\log\Delta)$ of them.
\end{proof}

Sampling a uniform \emph{vertex} of $X$ has two costs. For $X=P$ a uniform vertex is a uniform point,
$O(\log\Delta)$ counts. For a support $X=Q$ a uniform support cell is drawn by rejection
(\cref{sec:support-oracle}): sample a uniform point of $P$, accept its $h$-cell with probability
$1/\abs{P\cap C}$, which accepts with probability $K/n$ and so costs $\Otil(n/K)$ counts per support
vertex. We use the first in the bulk sampler for $P$ and the second in the sampler for $Q$.

\subsection{The spanner realized by range counts}\label{sec:wspd-spanner}

\paragraph{Representatives.}
For a nonempty cell $c$, its representative $\rep(c)$ is the vertex of $X$ inside $c$ that is first in the
$z$-order (the order a quadtree traversal visits leaves). For $X=P$ it is found by descending $\mathcal T$
from $c$, at each level entering the $z$-first nonempty child (one emptiness count per child), reaching a
single point in $O(\log\Delta)$ counts; for $X=Q$ the descent is over $h$-cells and returns the $z$-first
nonempty $h$-cell center, again $O(\log\Delta)$ counts. The $z$-order is a total order, so $\rep(c)$ is
unique and every search returns the same representative for the same cell; this fixes all ties. The
computation returns the \emph{coordinates} of $\rep(c)$, so exact Euclidean distances between
representatives are available with no further queries.

\paragraph{A locally testable pair family.}
A pair of cells $(c,c')$ is \emph{well-separated} if $\max\{r(c),r(c')\}\le\rho\,d(c,c')$, equivalently
$d(c,c')\ge\max\{r(c),r(c')\}/\rho$. The Callahan--Kosaraju recursion~\cite{CK95} on $\mathcal T$ produces a
well-separated pair decomposition $\mathcal W$ of $X$, but its membership is a property of the whole
recursion, not of a pair in isolation, so we do not enumerate $\mathcal W$ directly. Instead we enumerate a
\emph{larger}, locally testable family $\widehat{\mathcal W}$ and show that the spanner built from it
contains a standard WSPD spanner; the extra pairs are harmless. Write $p(c)$ for the quadtree parent of a
cell $c$, and call a pair $(c,c')$ a \emph{parent-pair} of itself if one of its cells is replaced by its
parent, i.e.\ $(p(c),c')$ or $(c,p(c'))$. Define the symmetric, unordered family
\[
  \widehat{\mathcal W}=\bigl\{\{c,c'\}:\ \text{$c,c'$ nonempty},\ \tfrac12\le r(c)/r(c')\le 2,\
   \text{$(c,c')$ well separated},\ \text{at least one parent-pair is \emph{not} well separated}\bigr\}.
\]
Membership $\{c,c'\}\in\widehat{\mathcal W}$ is decided in $O(1)$ counts: the size and separation tests are
geometric, the two parent-pairs are tested in $O(1)$ time, and the two emptiness tests are $O(1)$ counts.
Two facts about $\widehat{\mathcal W}$ are all we use.

\begin{enumerate}
\item \emph{$\widehat{\mathcal W}$ contains a WSPD.} Whatever orientation and tie-breaking the standard
recursion uses, a pair $(c,c')$ it outputs is well separated, has comparable cells $r(c')/2\le r(c)\le
2r(c')$ by~\cite[Lem.~22]{DMOSW25}, and was reached by one final subdivision of its immediate predecessor
in the recursion; that predecessor is exactly one of the parent-pairs $(p(c),c')$ or $(c,p(c'))$, and it
is \emph{not} well separated (else the recursion would have output it and stopped before subdividing).
Hence every pair of $\mathcal W$ satisfies the symmetric predicate, so $\mathcal W\subseteq\widehat{\mathcal
W}$, and this holds for either split side and for equal-size ties. (The converse fails -- the recursion may
stop at a well-separated ancestor pair and never visit a qualifying descendant -- so we claim only
containment, never equality.)
\item \emph{Bounded partner count.} Fix a cell $c$ of side $s$. Any $\widehat{\mathcal W}$-partner $c'$ has
$r(c')\in[s/2,2s]$, and some parent-pair is not well separated: if $(c,p(c'))$ is not well separated then
$d(c,p(c'))<r(p(c'))/\rho=O(s/\rho)$, and if $(p(c),c')$ is not well separated then $d(p(c),c')<
r(p(c))/\rho=O(s/\rho)$; either way $d(c,c')=O(s/\rho)$. The cells of side in $[s/2,2s]$ within
center-distance $O(s/\rho)$ of $c$ number $O(\rho^{-2})$, so $c$ has $O(\rho^{-2})$ partners; a vertex is
the representative of $O(\log\Delta)$ ancestor cells, so it has $O(\rho^{-2}\log\Delta)$ incident edges in
the spanner below.
\end{enumerate}

The \emph{spanner} $S=(X,E(S))$ has, for each $(c,c')\in\widehat{\mathcal W}$, the edge
$\{\rep(c),\rep(c')\}$ of \emph{exact Euclidean length} $\norm{\rep(c)-\rep(c')}$; $E(S)$ is the \emph{set}
of these edges, so identical edges from different pairs collapse. Every edge of $S$ is a genuine pair of
points of $X$ at its true distance, so $S$ is a Euclidean subgraph of the complete graph on $X$, giving the
left inequality $c_X(t)\le c_S(t)$ for every threshold. Because $\mathcal W\subseteq\widehat{\mathcal W}$,
$S$ \emph{contains} the standard WSPD spanner $S_{\mathcal W}$ as a subgraph, and a supergraph of a
$(1+\rho)$-spanner is a $(1+\rho)$-spanner; this is what \cref{sec:dt} needs, and it does not depend on
$\widehat{\mathcal W}$ equaling $\mathcal W$.

\paragraph{Stretch.}
We take the geometric separation a suitable constant multiple of $1/\rho$, so the standard WSPD spanner has
stretch at most $1+\rho$ (downstream $\rho=\Theta(\xi)=\Theta(\eps)$).

\begin{lemma}[Spanner stretch \cite{HarPeled11,DMOSW25}]\label{lem:stretch}
$S$ is a $(1+\rho)$-spanner of $X$: for all $x,y\in X$ there is an $x$--$y$ path in $S$ of length at most
$(1+\rho)\norm{x-y}$. Consequently $\wt(\MST(S))\le(1+\rho)\,\wt(\MST(X))$, and for every threshold $t$,
\begin{equation}\label{eq:appA-stretch}
  c_X(t)\le c_S(t)\le c_X\!\bigl(t/(1+\rho)\bigr).
\end{equation}
\end{lemma}

\begin{proof}
Let $S_{\mathcal W}$ be the spanner built from the standard WSPD $\mathcal W$ in the same way (one edge
$\{\rep(c),\rep(c')\}$ per pair). It is a $(1+\rho)$-spanner of $X$ by the WSPD spanner
theorem~\cite[Thm.~on WSPD spanners]{HarPeled11}: for $x,y\in X$, induct on the rank of $\norm{x-y}$ among
pairwise distances; the unique WSPD pair $(c,c')$ separating $x,y$ contributes the edge $\{\rep(c),
\rep(c')\}$, and well-separation gives $\norm{x-\rep(c)},\norm{y-\rep(c')}\le 2\rho\norm{x-y}$, each a
strictly shorter distance, so by induction $S_{\mathcal W}$ has short $x$--$\rep(c)$ and $y$--$\rep(c')$
paths whose concatenation with the edge has length $\le(1+\rho)\norm{x-y}$. Since $\mathcal W\subseteq
\widehat{\mathcal W}$ we have $S_{\mathcal W}\subseteq S$, and a supergraph of a $(1+\rho)$-spanner is a
$(1+\rho)$-spanner, so $S$ is a $(1+\rho)$-spanner; this is the spanner property. The MST bound follows
because $\MST(S)$ is no heavier than $\MST(S_{\mathcal W})\le(1+\rho)\wt(\MST(X))$ (Lemma~23
of~\cite{DMOSW25}; a supergraph has a no-heavier MST). For \eqref{eq:appA-stretch}: every $S$-edge is a
true $X$-pair, so the $S$-edges of length $\le t$ form a subgraph of the $X$-pairs of length $\le t$,
giving $c_X(t)\le c_S(t)$; conversely, if $\norm{x-y}\le t/(1+\rho)$ then the spanner path between $x,y$ has
every edge of length $\le(1+\rho)\norm{x-y}\le t$, so $c_S(t)\le c_X(t/(1+\rho))$.
\end{proof}

Integrating \eqref{eq:appA-stretch} against \eqref{eq:cluster-integral} yields the additive spanner bias
$A_L(X)\le A_L(H_X)\le(1+\rho)A_L(X)$ of \eqref{eq:spanner-stretch}.

\subsection{Local navigability}\label{sec:wspd-nav}

The death-time search needs, from a vertex $u$ of $S$, all its incident spanner edges of length at most a
given $r$, with their exact weights. We obtain these by enumerating the pairs of $\widehat{\mathcal W}$
incident to $u$ directly, without the contracted-graph oracle of \cite[Lemma~24]{DMOSW25} (which navigates
a threshold graph after cell contraction and does not return a vertex's defining pairs).

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.85, every node/.style={font=\footnotesize}]
  % cell c containing u
  \draw[fill=blue!8] (0,0) rectangle (1,1);
  \fill[blue!70] (0.28,0.32) circle (1.6pt); \node[blue!70] at (0.28,0.62) {$u=\rep(c)$};
  \node at (0.5,-0.32) {$c$, side $s$};
  % annulus of well-separated partners
  \draw[gray!55,dashed] (0.5,0.5) circle (2.0);
  \draw[gray!55,dashed] (0.5,0.5) circle (3.4);
  \node[gray!65] at (3.6,2.7) {partners at $d=\Theta(s/\rho)$};
  % a few partner cells with representatives + spanner edges
  \foreach \cx/\cy in {2.7/0.2, 0.2/2.6, 2.4/2.3, -1.9/1.4} {
    \draw[fill=red!8] (\cx,\cy) rectangle (\cx+1,\cy+1);
    \fill[red!70] (\cx+0.55,\cy+0.5) circle (1.4pt);
    \draw[blue!50!black, thick] (0.28,0.32) -- (\cx+0.55,\cy+0.5);
  }
  \node[red!70] at (3.25,0.0) {$\rep(c')$};
\end{tikzpicture}
\caption{The incident-edge oracle (\cref{lem:incident}). For a cell $c$ of side $s$ with $\rep(c)=u$, its
$\widehat{\mathcal W}$-partners $c'$ lie in an annulus at center-distance $\Theta(s/\rho)$ and have
comparable side; there are $O(\rho^{-2})$ of them, enumerated from $c$'s coordinates. Each partner
contributes a spanner edge $\{u,\rep(c')\}$ of exact Euclidean length. Scanning $u$'s $O(\log\Delta)$
ancestor cells where $u$ is the representative finds all incident edges.}
\label{fig:wspd}
\end{figure}

\begin{lemma}[Vertex incident-edge oracle]\label{lem:incident}
For a vertex $u\in X$ and a length $r$, all $S$-edges incident to $u$ of length at most $r$, with their
exact weights, can be enumerated using $O(\rho^{-2}\log^2\Delta)$ range-counting queries (illustrated in
\cref{fig:wspd}); their number is $O(\rho^{-2}\log\Delta)$.
\end{lemma}

\begin{proof}
An edge $\{u,u'\}\in E(S)$ comes from a pair $\{c,c'\}\in\widehat{\mathcal W}$ with $\{\rep(c),\rep(c')\}=
\{u,u'\}$; say $\rep(c)=u$, so $u\in c$ and $u$ is the $z$-first vertex of $c$. Hence $c$ is one of the
$O(\log\Delta)$ quadtree ancestor cells of (the leaf of) $u$. We scan all $O(\log\Delta)$ ancestor cells;
for each, one representative computation ($O(\log\Delta)$ counts) decides whether $\rep(c)=u$, which both
finds the relevant ancestors and accounts for testing them. For a relevant $c$ of side $s=r(c)$, its
$\widehat{\mathcal W}$-partners $c'$ have side in $[s/2,2s]$, and their distance is bounded \emph{above} by
the predicate's non-well-separated parent-pair (\cref{sec:wspd-spanner}, fact~(2)): whichever of
$(c,p(c'))$ or $(p(c),c')$ is not well separated gives $d(c,c')=O(s/\rho)$. The cells of side in $[s/2,2s]$
within center-distance $O(s/\rho)$ of $c$ therefore number $O(\rho^{-2})$ and are enumerated from $c$'s
coordinates with no query. For each candidate $c'$ we test nonemptiness ($1$ count) and the (symmetric,
unordered) $\widehat{\mathcal W}$-predicate ($O(1)$ counts), compute $\rep(c')$ ($O(\log\Delta)$ counts),
and, when $\norm{u-\rep(c')}\le r$, emit $\{u,\rep(c')\}$ with that exact weight. Each ancestor thus costs
$O(\rho^{-2})$ candidates at $O(\log\Delta)$ counts each, and there are $O(\log\Delta)$ ancestors, for
$O(\rho^{-2}\log^2\Delta)$ counts; the emitted edge set is deduplicated by vertex identity. Every incident
edge of length $\le r$ is found, since its defining $\widehat{\mathcal W}$-pair has $u$ as the
representative of one of its two cells, an ancestor of $u$ scanned above; only genuine edges are emitted,
each checked against the predicate and the exact length. The distinct-edge count $O(\rho^{-2}\log\Delta)$ is
the partner bound above.
\end{proof}

\subsection{The exact bottleneck death-time search}\label{sec:wspd-search}

We now compute the clipped death time $X_L(v)=\min\{\tau(v),L\}$ of a seed $v$ exactly. Fix once a
\emph{root} $r_X=\rep(\text{root cell of }\mathcal T)$, the global $z$-first vertex of $X$, and assign it
rank $-\infty$; assign every other vertex an independent uniform rank in $[0,1]$, drawn lazily and keyed
by \emph{vertex identity} (the point coordinates for $X=P$, the $h$-cell identifier for $X=Q$), so a
vertex has one rank no matter through which cell it is reached, and ranks are a.s.\ distinct (ties have
probability zero and are broken by $z$-order). As in \cref{sec:dt-sampler}, $\tau(v)$ is the bottleneck
(min--max-edge) distance from $v$ to the root or to any lower-ranked vertex.

\begin{algorithm}[t]
\caption{Clipped death time $X_L(v)$ on the WSPD spanner $S$}\label{alg:deathtime}
\KwIn{seed $v$; clip level $L$; ranks (root $r_X$ has rank $-\infty$)}
\KwOut{$\min\{\tau(v),L\}$}
$\beta(v)\leftarrow 0$; push $v$ on a min-priority queue keyed by $\beta$\;
\While{the queue is nonempty}{
  extract the minimum-$\beta$ vertex $u$ (ties broken by vertex identity, independent of ranks)\;
  \lIf{$u=r_X$ or $\mathrm{rank}(u)<\mathrm{rank}(v)$}{\Return $\beta(u)$}
  enumerate all incident edges $\{u,u'\}$ of weight $\le L$ (\cref{lem:incident})\;
  \lForEach{such edge of weight $\omega$}{$\beta(u')\leftarrow\min\{\beta(u'),\max\{\beta(u),\omega\}\}$}
}
\Return $L$\tcp*{queue emptied: $\tau(v)>L$}
\end{algorithm}

\paragraph{The search.}
Run a bottleneck variant of Dijkstra from $v$ (\cref{alg:deathtime}): maintain, for each reached vertex,
the smallest value $\beta(\cdot)$ such that $v$ reaches it by a path whose every edge is $\le\beta(\cdot)$,
in a priority queue keyed by $\beta$ over \emph{exact} edge weights. Repeatedly extract the queue-minimum vertex $u$.
If $u=r_X$ or $\mathrm{rank}(u)<\mathrm{rank}(v)$, output $\tau(v)=\beta(u)$ and stop. Otherwise expand
$u$ by enumerating, via \cref{lem:incident} with length bound $L$, \emph{all} its incident edges of weight
at most $L$, and relax each edge $\{u,u'\}$ of weight $\omega$ by
$\beta(u')\leftarrow\min\{\beta(u'),\max\{\beta(u),\omega\}\}$. Since only edges of weight at most $L$ are
relaxed, if the queue empties before any halt then $\tau(v)>L$, and we output $X_L(v)=L$. Expanding every extracted vertex over its full incident-edge set
(capped only at $L$) is exact bottleneck Dijkstra restricted to edges of weight $\le L$: the queue
extracts vertices in nondecreasing $\beta$, and when a vertex is extracted its $\beta$ is final because
any cheaper path would have been relaxed from an earlier extraction. Hence the value at the first
extraction of a root or lower-ranked vertex is exactly $\tau(v)$ when $\tau(v)\le L$, and the cutoff
returns exactly $\min\{\tau(v),L\}$; no rounding enters.

\paragraph{Per-search cost in expectation.}
Order the vertices by bottleneck distance from $v$, ties broken by vertex identity; this order is fixed
before the ranks are revealed and is the order in which the queue extracts them. Let $v=w_1,w_2,\dots$ be this order. The search halts at the first $w_i$ ($i\ge2$) that is the root or has
rank below $\mathrm{rank}(v)$, so it extracts $w_j$ only if none of $w_1,\dots,w_{j-1}$ is the root and $v$
has the smallest rank among the $j-1$ vertices $w_1,\dots,w_{j-1}$ (which include $v$). The latter has
probability $1/(j-1)$ for $j\ge2$, the $j-1$ ranks being independent and uniform, hence exchangeable. The expected number of extractions is thus at most
$1+\sum_{j\ge2}1/(j-1)=1+H_{\abs V-1}=O(\log\abs V)$. Each extraction expands one vertex by
\cref{lem:incident} at $O(\rho^{-2}\log^2\Delta)$ counts, so a search costs
$O(\rho^{-2}\log^2\Delta)\cdot O(\log\abs V)=\poly(\rho^{-1})\poly\log(n,\Delta)$ range counts in
expectation, independent of $n^{1/3}$, i.e.\ $\Otil_\rho(1)$.

\subsection{The global query budget}\label{sec:wspd-cap}

We do \emph{not} cap individual searches or impute a value to an unfinished one, since imputing $X_L=L$
would bias the death-time mean upward when $\tau(v)<L$. Instead we bound the whole estimator call.
The bulk sampler for $A_L(P)$ runs $s_P=O(\xi^{-2}n^{1/3})$ searches, each $\Otil_\rho(1)$ counts in
expectation (seeds drawn at $O(\log\Delta)$ counts by \cref{lem:telescoping}); the sampler for $A_L(Q)$
runs $s_Q=O_\xi(1)$ searches with seeds at $\Otil(n/K)=\Otil(n^{1/3})$ counts each. Hence the call has
expected query cost
\begin{equation}\label{eq:appA-totalcost}
  \overline Q=n^{1/3}\cdot\poly(\rho^{-1},\xi^{-1})\cdot\poly\log(n,\Delta)=\Otil_\eps(n^{1/3}).
\end{equation}
Fix a deterministic budget $B=C\overline Q$ for an absolute constant $C$ and abort the call, returning
\textsc{fail}, the moment the running count exceeds $B$. By Markov's inequality the call exceeds $B$ with
probability at most $1/C$, a small constant, and \textsc{fail} is treated by the geometric $W$-search of
\cref{sec:main-combine} as a bad run inside its median amplification. The uncapped sample average is the
exact clipped death-time average of \cref{sec:dt-sampler} and is unbiased; a run is counted bad if it
aborts or is inaccurate, and \cref{sec:main-combine} union-bounds the bad probability instead of
conditioning on non-abort. The cost is
$B=\Otil_\eps(n^{1/3})$ deterministically, and the displayed $\poly(1/\eps)$ factors in
\eqref{eq:appA-totalcost} are hidden by $\Otil_\eps$ for fixed $\eps$; for variable $\eps$ the bound reads
$n^{1/3}\poly(1/\eps)\poly\log(n,\Delta)$.

\paragraph{Summary.}
Collecting the four layers: a seed and any representative cost $O(\log\Delta)$ counts; the spanner is
defined and navigated locally, with each vertex expansion $O(\rho^{-2}\log^2\Delta)$ counts and each
search $\Otil_\rho(1)$ counts in expectation over $O(\log\abs V)$ expansions; the bulk sampler runs
$O(\xi^{-2}n^{1/3})$ searches and the support sampler $O_\xi(1)$ searches at $\Otil(n/K)$ each, all
governed by one global budget that aborts rarely. No step reveals a point beyond what a representative
computation already costs, no step uses a non-rectangular query, and the returned death-time averages are
exactly the unbiased clipped statistics of \cref{sec:dt-sampler}. This discharges the realization of the
locally navigable $(1+\rho)$-spanner deferred from \cref{sec:dt}, and with it the death-time primitive
runs within the $\Otil_\eps(n^{1/3})$ budget claimed in \cref{sec:main}.

\begin{remark}[Determinism of the construction]\label{rem:wspd-det}
The only randomness in this appendix is the rank assignment used by the death-time search; everything else
is a deterministic function of the input and the query history. In particular the representative of a cell,
the membership of a pair in $\widehat{\mathcal W}$, and the spanner edge set are fixed once and for all, so two searches
that touch the same cell see the same representative and the same incident edges. This consistency is what
makes the spanner a single well-defined graph $S$ and lets us reason about its death times as if $S$ were
written down, even though it is only ever probed one neighborhood at a time. The ranks may be drawn lazily
and reproducibly by hashing the vertex identity, so a vertex reached through two different cells in two
different searches still carries one rank; no global random tape is materialized.
\end{remark}

\begin{remark}[What a human verifier should check first]\label{rem:verify}
The two places where the implementation is most easily mis-stated, and which a careful reading should
target, are the local pair family (\cref{sec:wspd-spanner}) and the budget mechanism (\cref{sec:wspd-cap}).
For the former, the key point is that we do \emph{not} claim the locally testable family $\widehat{\mathcal
W}$ equals the recursion-generated WSPD $\mathcal W$; we claim only $\mathcal W\subseteq\widehat{\mathcal
W}$, so the spanner $S$ \emph{contains} a standard WSPD spanner (hence has stretch $1+\rho$) while every
extra edge of $S$ is still a true Euclidean pair (hence $c_X(t)\le c_S(t)$ is unharmed). The partner-count
bound, and so the per-vertex degree, comes from the predecessor pair being not well separated, not from
well-separation itself. For the latter, the point is that capping is global, not per-search, so that no individual
death-time value is ever imputed or conditioned on, which is what keeps the averages unbiased. These two
points are the substance of realizing the source's WSPD machinery under a counting oracle.

A third, more mechanical item worth checking is the support case $X=Q$ throughout: every operation that is
stated for points of $P$ has a parallel for cell centers, and the only difference is the cost of a uniform
seed, $O(\log\Delta)$ for $P$ versus $\Otil(n/K)$ for $Q$ by rejection. The representative of a support
cell is the $z$-first occupied $h$-subcell center, found by an emptiness descent that is identical in form
to the descent for $P$; the WSPD, the navigability, and the bottleneck search then run on $Q$ unchanged,
because a support $h$-cell is a quadtree cell and every region they query is a union of such cells. We have
verified that no step of the search ever needs a point of $P$ once the support is fixed, so the two
death-time samplers, on $P$ and on $Q$, are the same procedure run on two different vertex sets.
\end{remark}
