\section{Full proof of the support-MST estimator}\label{sec:app-lemma3}

This section completes the proof of \cref{lem:support}: the active-cover packing bound
(\cref{sec:app-packing}), the interleaving inequality (\cref{sec:app-interleave}), the two-part query
accounting that is the crux of the $\Otil(\sqrt K)$ bound (\cref{sec:app-accounting}), the accuracy of one
additive call (\cref{sec:app-accuracy}), and the guess-removal and approximate-$K$ steps
(\cref{sec:app-guess}). Throughout, $\beta\in(0,\tfrac1{16})$, $\sigma=\beta/32$, and $G\ge\Wq$ is the
current guess.

\subsection{The packing bound (\texorpdfstring{\cref{lem:packing}}{packing})}\label{sec:app-packing}

\begin{proof}[Proof of \cref{lem:packing}]
The bound has a lower half (spread points force MST weight up) and an upper half (no subset of $Q$ has MST
weight much above $\Wq$), and we combine them.

\emph{A spread subset.} The active cover $\mathcal B$ consists of $b$ pairwise-disjoint $\delta$-cells,
each nonempty in $Q$; index a cell by its integer grid coordinates $(i,j)$, so cell $(i,j)$ is the square
$[i\delta,(i{+}1)\delta)\times[j\delta,(j{+}1)\delta)$. Pick one support point of $Q$ from each block, and
color the block $(i,j)$ by $(i\bmod 2,\,j\bmod 2)\in\{0,1\}^2$, a four-coloring. If two blocks share a
color then their indices agree in parity, so they differ by at least $2$ in whichever coordinate they
differ; hence the two chosen points, lying in those cells, differ by at least $2\delta-\delta=\delta$ in
that coordinate and are at Euclidean distance $\ge\delta$. By pigeonhole some color class $S$ contains at
least $b/4$ of the chosen points, and they are pairwise at distance $\ge\delta$.

\emph{Spread points have heavy MST.} For a point set whose pairwise distances are all $\ge\delta$, every
spanning tree has each of its $\abs S-1$ edges of length $\ge\delta$, so $\wt(\MST(S))\ge\delta(\abs S-1)
\ge\delta(b/4-1)$.

\emph{The upper half.} We claim that for \emph{every} $S\subseteq Q$,
\begin{equation}\label{eq:doubling}
  \wt(\MST(S))\le 2\,\Wq .
\end{equation}
Let $T$ be the minimal subtree of $\MST(Q)$ that spans $S$ (the union of the $\MST(Q)$-paths between the
points of $S$); then $\wt(T)\le\wt(\MST(Q))=\Wq$. Double every edge of $T$ to obtain a connected multigraph
in which every vertex has even degree; it therefore has an Euler tour, a closed walk of length $2\wt(T)$
visiting every vertex of $T$ (hence every point of $S$) at least once. Traverse the tour and shortcut past
already-visited vertices: by the triangle inequality in the Euclidean plane each shortcut does not
increase the length, so we obtain a closed tour through exactly the points of $S$ of length $\le 2\wt(T)$.
Deleting its longest edge leaves a spanning path of $S$, and $\MST(S)$ is no heavier than this path, so
$\wt(\MST(S))\le 2\wt(T)\le 2\Wq$, which is \eqref{eq:doubling}.

\emph{Combining.} Taking $S$ to be the spread color class, $2\Wq\ge\wt(\MST(S))\ge\delta(b/4-1)$, i.e.\
$\Wq\ge\delta(b-4)/8$. For $b\ge 8$ the right side is at least $\delta(b-b/2)/8=b\delta/16$, which is the
claim. If $b<8$ then $K=O(1)$: the support has only a constant number of cells, found explicitly by the
$\Otil(\sqrt K)$ cover construction, and $\Wq$ is then computed exactly from those $O(1)$ points, so the
inequality is not needed.
\end{proof}

\subsection{The filtration (\texorpdfstring{\eqref{eq:interleave}}{interleaving})}\label{sec:app-interleave}

\begin{proof}[Proof of \eqref{eq:interleave}]
Fix $r$ and write $a=a(r)$. If $a=h$ then $\Gamma_r$ is the Euclidean threshold graph on the support
points themselves, so $\cG(r)=\cQ(r)$ and both inequalities hold. Assume $a>h$. For the upper bound, every
edge $\{x,y\}$ of the Euclidean $r$-graph on $Q$ has its endpoints in occupied cells at set-distance
$\le r$, which are therefore adjacent (or equal) in $\Gamma_r$; contracting each cell to a vertex only
merges components, so $\cG(r)\le\cQ(r)$. For the lower bound, if two cells are adjacent in $\Gamma_r$ then
any two points inside them are at distance at most $r+2\sqrt2\,a$, and two points in one cell at most
$2\sqrt2\,a\le r+2\sqrt2\,a$; since $a\le\sigma r/64$ gives $2\sqrt2\,a<\sigma r$, every $\Gamma_r$-edge
joins points within $r^+=(1+\sigma)r$, so components of $\Gamma_r$ refine those of the Euclidean
$r^+$-graph, i.e.\ $\cQ(r^+)\le\cG(r)$.
\end{proof}

By \eqref{eq:components-edges}, $\cQ(t)-1=\abs{\{e\in\MST(Q):\abs e>t\}}\le\Wq/t$, so by \eqref{eq:interleave}
\begin{equation}\label{eq:U-bound}
  \cG(r)\le\cQ(r)\le 1+\Wq/r\le 1+G/r=U(r).
\end{equation}
This is the bound on the \emph{component} count used by the leader estimator.

\subsection{Query accounting: both sums are \texorpdfstring{$\Otil(\sqrt K)$}{O(sqrt K)}}\label{sec:app-accounting}

This is the heart of \cref{lem:support}, and the place the separation of the two cost drivers pays off:
the number of trials is set by the variance, which sees only the component count, while the work per trial
sees the occupied-cell count, and the two are summed against different parts of the packing bound so that
neither alone forces $\Omega(K)$.

\paragraph{The cost model.} One additive call runs the leader estimator at each fine scale $r_j$ (those
with $a(r_j)\le\delta$, $j=0,\dots,T-1$) with candidate universe of size $M_j=M(r_j)$ from
\eqref{eq:Msize}, target additive accuracy $\alpha_j=\beta G/(32\,T\sigma r_j)$, and a-priori bound
$U_j=1+G/r_j\ge c_j$ on the component count (from \eqref{eq:U-bound}). Two costs arise at scale $j$.

\emph{Trials.} By \cref{lem:leader} a single trial has variance $\Var(Z)\le M_jc_j\le M_jU_j$. To estimate
$c_j$ to additive error $\alpha_j$ with failure probability $\delta'=1/(6T)$, the median-of-means recipe
of \cref{sec:prelim} averages $O(\Var/\alpha_j^2)$ trials per block and takes a median of $O(\log(1/\delta'))
=O(\Lambda)$ blocks, for a total of
\begin{equation}\label{eq:kj}
  k_j=O\!\Bigl(\frac{M_jU_j}{\alpha_j^2}\,\Lambda\Bigr),\qquad \Lambda=\log\frac{n\Delta}{h\eta}
\end{equation}
trials. Each trial spends one range count to test the sampled cell for occupancy, so the trials alone
cost $\sum_j k_j$ range counts.

\emph{Explorations.} A trial does extra work only when its sampled cell is occupied, which happens with
probability $N_j/M_j$ (where $N_j$ is the number of occupied $a(r_j)$-cells), and an occupied exploration
examines $O(\log K)$ cells in expectation (the harmonic bound of \cref{sec:support-estimator}), each at
$O(\sigma^{-2})$ counts to find its neighbors. So the expected exploration cost at scale $j$ is
$O(k_j\cdot(N_j/M_j)\cdot\log K\cdot\sigma^{-2})=O\bigl((N_jU_j/\alpha_j^2)\,\Lambda\log K\,\sigma^{-2}\bigr)$,
in which the universe size $M_j$ has cancelled and the occupied-cell count $N_j$ has taken its place.

We bound the two scale-sums $\sum_j M_jU_j/\alpha_j^2$ and $\sum_j N_jU_j/\alpha_j^2$ separately; the
remaining $\Lambda,\log K,\sigma^{-2}$ factors are polylogarithmic and tracked at the end.

\paragraph{Candidate-root sum (uses $U_j\ge c_j$).}
Substituting \eqref{eq:Msize} and $\alpha_j=\beta G/(32T\sigma r_j)$,
\begin{align}
  \frac{M_jU_j}{\alpha_j^2}
   &\le C\,\frac{b\delta^2}{\sigma^2 r_j^2}\Bigl(1+\frac G{r_j}\Bigr)\frac{T^2\sigma^2 r_j^2}{\beta^2G^2}
   = C\,\frac{T^2}{\beta^2}\Bigl(\frac{b\delta^2}{G^2}+\frac{b\delta^2}{G r_j}\Bigr).
   \label{eq:cand-per-scale}
\end{align}
We now bound the two terms of \eqref{eq:cand-per-scale} using the packing bound $G\ge\Wq\ge b\delta/16$
(\cref{lem:packing}). For the first term, $b\delta^2/G^2=b(\delta/G)^2\le b(16/b)^2=256/b$ is
\emph{scale-independent}, so summed over the $T$ scales it contributes $O(T/b)$. For the second term,
$b\delta^2/(Gr_j)=(b\delta/G)(\delta/r_j)\le 16\,\delta/r_j$, and the thresholds $r_j=r_0(1+\sigma)^j$ form
a geometric progression, so
\[
  \sum_{j\ge0}\frac1{r_j}=\frac1{r_0}\sum_{j\ge0}(1+\sigma)^{-j}=\frac1{r_0}\cdot\frac{1+\sigma}{\sigma}
   =O\!\Bigl(\frac1{\sigma r_0}\Bigr).
\]
Since $r_0=\max\{h,\beta G/(32K)\}\ge\beta G/(32K)\ge\beta(b\delta/16)/(32K)=\beta b\delta/(512K)$, we get
$\delta/r_0\le 512K/(\beta b)=O(K/(\beta b))$, so the second term sums to $O(\delta/(\sigma r_0))=
O(K/(\sigma\beta b))$. Adding the two and multiplying by the prefactor $CT^2/\beta^2$ from
\eqref{eq:cand-per-scale}, and using $T/b=O(K/(\sigma\beta b))$ (as $T=\Otil(1)$ and $K\ge\sigma\beta T$
for large $K$),
\begin{equation}\label{eq:cand-total}
  \sum_j \frac{M_jU_j}{\alpha_j^2}
   = O\!\Bigl(\frac{T^2}{\beta^2}\Bigl(\frac{T}{b}+\frac{1}{\sigma}\cdot\frac{\delta}{r_0}\Bigr)\Bigr)
   = O\!\Bigl(\frac{T^2}{\beta^2}\cdot\frac{K}{\sigma\beta b}\Bigr)
   = \Otil_\beta\!\Bigl(\frac Kb\Bigr)=\Otil_\beta(\sqrt K),
\end{equation}
using $b=\Theta(\sqrt K)$ from \eqref{eq:cover-size}. Multiplying by $\Lambda$ keeps this
$\Otil_\beta(\sqrt K)$.

\paragraph{Exploration sum (uses $N_j\le K$).}
Here the occupied-cell count enters, but only through $N_j\le K$:
\begin{equation}\label{eq:expl-per-scale}
  \frac{N_jU_j}{\alpha_j^2}\le K\Bigl(1+\frac G{r_j}\Bigr)\frac{T^2\sigma^2 r_j^2}{\beta^2 G^2}
   = \frac{KT^2\sigma^2}{\beta^2}\Bigl(\frac{r_j^2}{G^2}+\frac{r_j}{G}\Bigr).
\end{equation}
The difference from the candidate-root sum is that the exploration sum runs only over \emph{fine} scales. A scale contributes explorations only when $a(r_j)\le\delta$, and the floor rule for
$a(\cdot)$ (\cref{sec:support-filtration}) makes $a(r_j)=\Theta(\sigma r_j)$ unless it has hit its floor;
either way $a(r_j)\le\delta$ forces $r_j\le R\eqdef 128\delta/\sigma$. So the fine scales are a
$(1+\sigma)$-geometric sequence \emph{capped above} at $R$, and geometric sums are dominated by their
largest term:
\[
  \sum_{\text{fine}}r_j\le R\sum_{i\ge0}(1+\sigma)^{-i}=O\!\Bigl(\frac R\sigma\Bigr)=O\!\Bigl(\frac
   {\delta}{\sigma^2}\Bigr),\qquad
  \sum_{\text{fine}}r_j^2\le R^2\sum_{i\ge0}(1+\sigma)^{-2i}=O\!\Bigl(\frac{R^2}\sigma\Bigr)=O\!\Bigl(\frac
   {\delta^2}{\sigma^3}\Bigr).
\]
Substituting these into the two terms of \eqref{eq:expl-per-scale},
\begin{equation}\label{eq:expl-total}
  \sum_{\text{fine }j}\frac{N_jU_j}{\alpha_j^2}
   \le\frac{KT^2\sigma^2}{\beta^2}\Bigl(\frac1{G^2}\sum r_j^2+\frac1G\sum r_j\Bigr)
   = O\!\Bigl(\frac{KT^2}{\beta^2}\Bigl(\frac{\delta^2}{\sigma G^2}+\frac{\delta}{G}\Bigr)\Bigr).
\end{equation}
Now the packing bound $G\ge\Wq\ge b\delta/16$ and $b=\Theta(\sqrt K)$ tame the two $K$-bearing factors:
$K\delta/G\le 16K/b=O(K/\sqrt K)=O(\sqrt K)$, and $K\delta^2/G^2\le 256K/b^2=O(K/K)=O(1)$. Hence
\eqref{eq:expl-total} is $\Otil_\beta(\sqrt K)$, the second occupied-cell branch matching the first.
A predetermined global cap of $\Theta(\Lambda\log K)\cdot\Otil_\beta(\sqrt K)$ expansions, enforced by
aborting the call when it is exceeded, converts this expectation into a worst-case query budget: by Markov
the cap is exceeded with probability $O(1)$, which the median amplification of \cref{sec:support-call}
absorbs.

Adding the two branches and the $\Otil(\sqrt K)$ emptiness queries of the cover, one additive call uses
$\Otil_\beta(\sqrt K)$ range-counting queries; tracing $\sigma=\beta/32$, $T=O(\beta^{-1}\log(\Delta/h))$,
degree $O(\beta^{-2})$, and the amplification gives the explicit $O(\beta^{-8}\sqrt K\,\Lambda^{O(1)})$.

\subsection{Accuracy of one additive call (\texorpdfstring{\cref{lem:additive}}{additive})}\label{sec:app-accuracy}

\begin{proof}[Proof of \cref{lem:additive}]
By the cluster-count integral \eqref{eq:cluster-integral}, $\Wq=\int_0^\infty(\cQ(t)-1)\,dt$. We compare
the estimator \eqref{eq:additive-call} to this integral and bound the discrepancy by three terms, each at
most $\beta G/32$: the part of the integral below $r_0$, the Riemann discretization gap, and the
statistical error. We then sum them.

\emph{The interval below $r_0$.} Distinct support centers lie at distance $\ge h$, so $\cQ(t)=K$ for $t<h$
and
\[
  \int_0^h(\cQ(t)-1)\,dt=(K-1)h,
\]
the leading $(K-1)h$ term of \eqref{eq:additive-call}. If $r_0>h$, the integral over $[h,r_0]$ is omitted
by the estimator; since $\cQ(t)-1\le K-1<K$ there, it contributes at most $\int_h^{r_0}K\,dt\le Kr_0$, and
the choice $r_0=\beta G/(32K)$ (when it exceeds $h$) makes this $\le\beta G/32$. If instead $r_0=h$ this
term is absent.

\emph{The Riemann gap.} On $[r_0,\infty)$ the estimator forms $\sum_j d_j(c_j-1)$ with $d_j=\sigma r_j$ the
width of the $j$-th step and $c_j=\cG(r_j)$ the true scale-graph component count. By the interleaving
\eqref{eq:interleave}, $\cQ(r_{j+1})\le c_j\le\cQ(r_j)$, so $d_j(c_j-1)$ lies between the lower and upper
Riemann rectangles of $\int_{r_j}^{r_{j+1}}(\cQ(t)-1)\,dt$ on that step. Hence $\sum_j d_j(c_j-1)$ and the
integral differ by at most the total staircase drop,
\[
  \sum_j d_j\bigl(\cQ(r_j)-\cQ(r_{j+1})\bigr).
\]
An MST edge of $Q$ of length $\ell$ lies in a unique band $(r_j,r_{j+1}]$ and there decreases $\cQ$ by one,
contributing $d_j=\sigma r_j\le\sigma\ell$ to this sum and nothing elsewhere; summing over edges, the gap
is at most $\sigma\sum_{e\in\MST(Q)}\abs e=\sigma\Wq$. With $\sigma=\beta/32$ this is $\le\beta\Wq/32\le
\beta G/32$ (using $G\ge\Wq$).

\emph{The statistical error.} The estimator uses $\widehat c_j$ in place of $c_j$, so its deviation from
the exact discretized sum is $\sum_j d_j\abs{\widehat c_j-c_j}$. On the good event $\abs{\widehat c_j-c_j}
\le\alpha_j$ for all $j$, and substituting $d_j=\sigma r_j$ and $\alpha_j=\beta G/(32\,T\sigma r_j)$,
\[
  \sum_j d_j\abs{\widehat c_j-c_j}\le\sum_{j=0}^{T-1}\sigma r_j\cdot\frac{\beta G}{32\,T\sigma r_j}
   =\sum_{j=0}^{T-1}\frac{\beta G}{32\,T}=\frac{\beta G}{32},
\]
the per-scale accuracies having been chosen precisely so that the widths $d_j$ cancel and the $T$ terms
sum to $\beta G/32$.

\emph{Summing.} Adding the three contributions, $\abs{\widehat W(G)-\Wq}\le 3\beta G/32<\beta G$, which is
the accuracy claim. The good event requires all $T$ scale-estimates to be within their $\alpha_j$
simultaneously; amplifying each to failure probability $\le 1/(6T)$ by a median over $O(\log T)$ blocks
and union-bounding gives the good event probability $\ge 5/6$, the $O(\log T)=O(\Lambda)$ overhead already
included in the cost of \cref{sec:app-accounting}.
\end{proof}

\subsection{Removing the guess and approximate \texorpdfstring{$K$}{K}}\label{sec:app-guess}

Run \cref{lem:additive} with $\beta_0=\tfrac1{32}$ on the decreasing guesses $G=U_Q,U_Q/2,\dots$, where
$U_Q=(K-1)\diam(Q)\ge\Wq$, amplified so that with constant probability every call with $G\ge\Wq$ is
correct, and stop at the first $G$ with $\widehat W(G)\ge G/4$. If $G\ge8\Wq$ then $\widehat W(G)\le
\Wq+G/32\le G/8+G/32<G/4$, so no such $G$ stops the search; if $G\in[\Wq,2\Wq)$ then $\widehat W(G)\ge
\Wq-G/32>G/4$, so the search stops while $G\ge\Wq$. Hence the selected $G_0$ obeys $\Wq\le G_0<8\Wq$. A
final call with $\beta=\eta/16$ then gives $\abs{\widehat\Wq-\Wq}\le(\eta/16)G_0<(\eta/2)\Wq$. There are
$O(\log(\Delta/h))$ guesses, so the total cost is $\Otil_\eta(\sqrt K)$.

It remains to replace the exact $K$ by $\widehat K=(1\pm\kappa)K$ with $\kappa=\eta/1000$
(\cref{sec:support-oracle}). Then $q=\ceil{\smash{\sqrt{\widehat K}}}$ still gives $b=\Theta(\sqrt K)$; the
baseline term $(K-1)h$ changes by at most $\kappa K h\le\kappa\Wq$ (since $\Wq\ge(K-1)h$ by
\eqref{eq:cluster-integral}); $r_0$ changes by a $1+O(\kappa)$ factor; and every bound in
\cref{sec:app-accounting}--\cref{sec:app-accuracy} changes by a constant factor. The final error is thus
at most $\eta\Wq$. Estimating $K$ costs $\Otil_\eta(n/K)$ queries, the only place the $n/K$ term arises.
This completes the proof of \cref{lem:support}.\qed
