\section{Full proofs for the death-time primitive and the reduction}\label{sec:app-assembly}

\subsection{The death-time sampler}\label{sec:app-dt}

\begin{proof}[Proof of \eqref{eq:death-multiset} and \eqref{eq:dt-var}]
Fix a rank assignment and a threshold $t$, and let $G_H(t)$ be the graph on $V$ with the edges of $H$ of
weight $\le t$. A vertex $v$ has $\tau(v)>t$ iff every path in $H$ from $v$ to the root or to a
lower-ranked vertex uses an edge of weight $>t$; equivalently the component of $v$ in $G_H(t)$ contains
neither the root nor any vertex of smaller rank, i.e.\ $v$ is the minimum-rank vertex of a component of
$G_H(t)$ that does not contain the root. Each component of $G_H(t)$ has exactly one minimum-rank vertex,
and one of these components is the root's, so
\[
  \abs{\{v:\tau(v)>t\}}=(\#\text{components of }G_H(t))-1=c_H(t)-1,
\]
which by \eqref{eq:components-edges} equals the number of MST edges of weight $>t$. Two nonnegative
integer-valued measures on $[0,\infty)$ with equal right tails at every $t$ are equal, so the multiset
$\{\tau(v)\}$ equals $\{0\}\cup\{\wt(e):e\in\MST(H)\}$ (the lone $0$ being the root's death time); this is
\eqref{eq:death-multiset}.

For the clipped statistic, integrate the tail identity against the clip kernel. Since $\min\{\tau(v),L\}
=\int_0^L\mathbf 1[\tau(v)>t]\,dt$,
\[
  \sum_{v\in V}\min\{\tau(v),L\}=\int_0^L\abs{\{v:\tau(v)>t\}}\,dt=\int_0^L(c_H(t)-1)\,dt
    =\sum_{e\in\MST(H)}\min\{\wt(e),L\}=A_L(H),
\]
so for a uniformly random $v$ we have $\E X_L=A_L(H)/\abs V$, i.e.\ $\abs V\cdot\E X_L=A_L(H)$, and the
sample mean scaled by $\abs V$ is unbiased for $A_L(H)$. For the variance, $0\le X_L\le L$ gives
$X_L^2\le L\,X_L$, hence
\[
  \Var(X_L)\le\E X_L^2\le L\,\E X_L=\frac{L\,A_L(H)}{\abs V}\le\frac{L\,\wt(\MST(H))}{\abs V}.
\]
We average over $s$ seeds drawn independently and uniformly from $V$. A single run shares one rank
assignment across all seeds (the ranks are drawn once, lazily, keyed by vertex identity), so the
$X_L(v_i)$ are not unconditionally independent; we therefore condition on the entire rank assignment. For
any \emph{fixed} ranking, $X_L(\cdot)$ is a fixed function on $V$, the seeds $v_1,\dots,v_s$ are
independent and uniform, so the $X_L(v_i)$ are i.i.d., with mean $\E[X_L\mid\text{ranking}]=\tfrac1{\abs V}
\sum_{v}X_L(v)=A_L(H)/\abs V$ -- the \emph{same value for every ranking}, by the tail identity above which
holds rank-by-rank. Hence the conditional mean of $\tfrac{\abs V}{s}\sum_i X_L(v_i)$ is $A_L(H)$
regardless of the ranking (so the estimator is unbiased), and its conditional variance is
$\tfrac{\abs V^2}{s}\Var[X_L\mid\text{ranking}]\le\abs V\,L\,\wt(\MST(H))/s$ by the display above, again
uniformly over rankings; by the law of total variance the unconditional variance obeys the same bound.
This is \eqref{eq:dt-var}, and it needs no independence between seeds beyond conditioning on the shared
ranking.
\end{proof}

\begin{proof}[Proof of \eqref{eq:spanner-stretch}]
We first bracket the threshold-graph component counts, then integrate. Since every edge of $H_X$ is a true
pair of $X$ (\cref{lem:stretch}), the graph of $H_X$-edges of length $\le t$ is a subgraph of all $X$-pairs
of length $\le t$, so it has at least as many components: $c_{H_X}(t)\ge c_X(t)$. For the other direction,
$H_X$ is a $(1+\rho)$-spanner, so any two points $x,y\in X$ with $\norm{x-y}\le t/(1+\rho)$ are joined in
$H_X$ by a path whose every edge has length $\le(1+\rho)\norm{x-y}\le t$; hence $x,y$ lie in one component
of the $H_X$-edges of length $\le t$, giving $c_{H_X}(t)\le c_X(t/(1+\rho))$. Together,
\begin{equation}\label{eq:HX-bracket}
  c_X(t)\le c_{H_X}(t)\le c_X\!\bigl(t/(1+\rho)\bigr).
\end{equation}
Now integrate $c-1$ against the clip kernel $\mathbf 1[t\le L]$. The left inequality of \eqref{eq:HX-bracket}
gives directly $A_L(X)=\int_0^L(c_X-1)\le\int_0^L(c_{H_X}-1)=A_L(H_X)$. For the upper bound, substitute
$t=(1+\rho)s$ in $\int_0^L(c_{H_X}(t)-1)\,dt\le\int_0^L(c_X(t/(1+\rho))-1)\,dt$:
\[
  A_L(H_X)\le\int_0^{L}\!\bigl(c_X(t/(1+\rho))-1\bigr)\,dt=(1+\rho)\!\int_0^{L/(1+\rho)}\!\!(c_X(s)-1)\,ds
   \le(1+\rho)\,A_L(X),
\]
the last step because the integrand is nonnegative and $L/(1+\rho)\le L$. This is \eqref{eq:spanner-stretch};
the additive inflation is $A_L(H_X)-A_L(X)\le\rho\,A_L(X)\le\rho\,\wt(\MST(X))$.
\end{proof}

\subsection{Support regularization and snapping}\label{sec:app-snap}

\begin{proof}[Proof of \eqref{eq:Kh-packing}]
Choose one point of $P$ in each nonempty $h$-cell and four-color the cells by index parities. In each
color class the chosen points are pairwise at distance $\ge h$, so the class has $\le 1+\wt(\MST(\text{class}))/h
\le 1+2W/h$ points by \eqref{eq:doubling} applied to $P$. Summing the four classes,
$K_h\le 4+8W/h$.
\end{proof}

\begin{lemma}[The regularization search]\label{lem:kwindow}
Fix a small constant $c\in(0,1)$ and set $C_\xi=\Theta(1/\xi)$ with $C_\xi/c>16$. Given a guess $G\ge W$,
start at $h_0=\Theta(\xi L)$ and, if the nonempty-cell count $K_{h_0}$ is certified below $cK_0$, halve $h$
until it is first certified at least $cK_0$. The resulting scale $h$ satisfies $h\le h_0=\Theta(\xi L)$ and
$\tfrac{c}{4}K_0\le K=K_h\le C_\xi K_0$, using $\Otil_\xi(n/K_0)=\Otil(n^{1/3})$ range-counting queries in
total.
\end{lemma}

\begin{proof}
As $h$ halves, each nonempty cell splits into at most four, so $K_h$ is nondecreasing as $h$ decreases and
grows by a factor at most $4$ per halving. We only ever halve, so the final scale obeys $h\le
h_0=\Theta(\xi L)$, which is what the snapping bound $\delta_s=\sqrt2\,h=O(\xi L)$ of \cref{sec:dt-snap}
needs.

\emph{Upper bound on $K$.} At the initial scale, $K_{h_0}\le 8W/h_0+4=O(K_0/\xi)\le C_\xi K_0$ by
\eqref{eq:Kh-packing} and $h_0=\Theta(\xi G/K_0)\ge\Theta(\xi W/K_0)$. If $K_{h_0}\ge cK_0$ we stop at
$h_0$ and $K=K_{h_0}\in[cK_0,C_\xi K_0]$. Otherwise let $h$ be the first halved scale certified at least
$cK_0$; the previous scale was below $cK_0$, so the quadrupling bound gives $K<4cK_0\le C_\xi K_0$. (A
finest scale always crosses, since at cell side $1$ every point is its own cell and $K=n\ge K_0$.) Either
way $K\le C_\xi K_0$.

\emph{Lower bound on $K$ and the test.} At each scale we run a fixed batch of $m=\Theta(\xi^{-2}(n/K_0)\log
n)$ rejection trials (\cref{sec:support-oracle}): a trial draws a uniform point of $P$ ($O(\log\Delta)$
counts, \cref{lem:telescoping}) and accepts with probability $1/\abs{P\cap C}$, so it accepts with
probability exactly $K_h/n$. A Chernoff bound on the accept count separates $K_h\ge cK_0$ from
$K_h<\tfrac{c}{4}K_0$ with failure probability $n^{-\Omega(1)}$. Since there are only $O(\log\Delta)$
candidate scales, a union bound makes the threshold test ``accepts $\ge\tfrac{c}{2}K_0\,m/n$'' correct at
\emph{all} of them simultaneously with probability $1-n^{-\Omega(1)}$; on that good event the adaptively
chosen stopping scale -- the first the test accepts -- is also correct, so no separate argument for the
data-dependent stop is needed. There $K_h\ge\tfrac{c}{4}K_0$ (else the test would have rejected), and with
the upper bound $K\in[\tfrac{c}{4}K_0,C_\xi K_0]=\Theta_\xi(K_0)$, which is \eqref{eq:K-window} with the
constants relabeled.

\emph{Cost.} The test spends $\Otil(m)=\Otil_\xi(n/K_0)$ counts per scale regardless of how small $K_h$
is, and runs at $O(\log\Delta)$ scales, for $\Otil_\xi(n/K_0)=\Otil(n^{1/3})$ total. After $h$ is fixed we
estimate $K$ to relative error $O(\xi)$ with a \emph{fresh} inverse-binomial batch at scale $h$
(\cref{sec:support-oracle}), so the estimate is independent of the adaptive stopping rule; this costs a
further $\Otil_\xi(n/K)=\Otil_\xi(n^{1/3})$ and is the input to \cref{lem:support} required by
\cref{sec:main-pieces}. Both the support estimator
$\Otil_\eta(\sqrt K+n/K)$ and the support death-time sampler are run at this $K=\Theta_\xi(K_0)$, and
since $\sqrt{C_\xi K_0}=\Otil_\xi(n^{1/3})$ and $n/(\tfrac{c}{4}K_0)=\Otil_\xi(n^{1/3})$, all three pieces
of \cref{sec:main-pieces} remain $\Otil_\xi(n^{1/3})$.
\end{proof}

\begin{proof}[Proof of \eqref{eq:snap-interleave}--\eqref{eq:tail-transfer} and $\Wq\le C_W W$]
\emph{Interleaving.} Snapping moves each point by at most $\sqrt2\,h/2$ (the half-diagonal of an
$h$-cell), so for any two points the change in their distance is at most $2\cdot\sqrt2\,h/2=\sqrt2\,h$; we
write $\delta_s=\sqrt2\,h$, noting that the cleaner one-sided statements below are what we use. Concretely, if two original points are within
$t$ then their snapped centers are within $t+\delta_s$, and if two centers are within $t$ then the
original points are within $t+\delta_s$. The first statement says a $P$-component at threshold $t$ stays
connected in $Q$ at threshold $t+\delta_s$, so $\cQ(t+\delta_s)\le c_P(t)$; the second says a
$Q$-component at threshold $t$ lifts to a $P$-component at threshold $t+\delta_s$, so $c_P(t+\delta_s)\le
\cQ(t)$. Reindexing gives
\[
  c_P(t+\delta_s)\le\cQ(t)\le c_P(t-\delta_s)\qquad(t\ge\delta_s),
\]
which is \eqref{eq:snap-interleave}.

\emph{Tail transfer.} Write $B_L(X)=\int_L^\infty(c_X(t)-1)\,dt=\sum_{e\in\MST(X)}(\abs e-L)_+$. Integrating
the interleaving over $[L,\infty)$ and shifting the variable,
\[
  B_{L+\delta_s}(P)\le B_L(Q)\le B_{L-\delta_s}(P),
\]
so $B_L(Q)$ and $B_L(P)$ both lie in $[B_{L+\delta_s}(P),B_{L-\delta_s}(P)]$, an interval of width
\[
  B_{L-\delta_s}(P)-B_{L+\delta_s}(P)=\int_{L-\delta_s}^{L+\delta_s}(c_P(t)-1)\,dt .
\]
Bounding the integrand by Markov, $c_P(t)-1\le W/t$ \eqref{eq:components-edges}, and using $\delta_s\le
L/2$ so that $t\ge L-\delta_s\ge L/2$ on the range,
\[
  \abs{B_L(Q)-B_L(P)}\le\int_{L-\delta_s}^{L+\delta_s}\frac Wt\,dt\le\frac{2\delta_s}{L-\delta_s}\,W
   \le\frac{4\delta_s}{L}\,W=O(\xi W),
\]
since $\delta_s=\sqrt2\,h=O(\xi L)$ by \cref{lem:kwindow}; this is \eqref{eq:tail-transfer}.

\emph{Support weight.} Finally we bound $\Wq=\wt(\MST(Q))$. Map each point of $P$ to its $h$-cell center;
this sends the $n-1$ edges of $\MST(P)$ to a connected \emph{multigraph} on the $K$ support centers (it may
have parallel edges and loops). The image of an $\MST(P)$-edge has length at most its original length plus
$\delta_s$, since each of its two endpoints moves by at most $\sqrt2\,h/2=\delta_s/2$, so their distance
changes by at most $\delta_s$. Take any spanning tree of this multigraph: it has $K-1$ edges, each the
image of a distinct $\MST(P)$-edge, so its weight is at most $\sum_{e\in\MST(P)}\abs e+\delta_s(K-1)\le
W+\delta_s(K-1)$. Since $\MST(Q)$ is no heavier than this spanning tree, $\Wq\le W+\delta_s(K-1)$.
To finish, $\delta_s(K-1)\le\delta_s K=\sqrt2\,h\cdot K$; by the packing bound \eqref{eq:Kh-packing} and
the window $K=\Omega(K_0)$ of \cref{lem:kwindow}, $K\le 8W/h+4$ gives $h\le 8W/(K-4)=O(W/K_0)$, so
\[
  \delta_s K=\sqrt2\,h\,K\le\sqrt2\,h(8W/h+4)=8\sqrt2\,W+4\sqrt2\,h=O(W) .
\]
Hence $\Wq\le W+O(W)=C_W W$ for an absolute constant $C_W$.
\end{proof}

\subsection{The assembly bound (\texorpdfstring{\eqref{eq:guess-accuracy}}{guess accuracy})}\label{sec:app-combine}

\begin{proof}[Proof of \eqref{eq:guess-accuracy}]
Assume $G\ge W$, so $L=G/K_0\ge W/K_0$ and $\delta_s=O(\xi L)$ satisfies $\delta_s\le L/2$ for small
$\xi$. We give each of the four contributions to \eqref{eq:combine}--\eqref{eq:decomp} a budget of
$\tfrac{C_*}{4}\xi G$ and verify each in turn; set $\rho=\Theta(\xi)$ and recall $W\le G$ and $\Wq\le C_W
W\le C_W G$.

\emph{(i) Bulk of $P$, $\widehat A_P$.} Two errors enter. The spanner bias is one-sided and at most $\rho
W$ by \eqref{eq:spanner-stretch}, which is $\le\tfrac{C_*}{12}\xi G$ for $\rho$ a small enough multiple of
$\xi$. The sampling error: by \eqref{eq:dt-var} with $\abs V=n$, $L=G/K_0$, and $\wt(\MST(H_P))\le(1+\rho)W
\le 2G$, the variance of the $s_P$-sample mean (scaled by $n$) is at most $nL\cdot 2G/s_P=2nG^2/(K_0 s_P)$;
with $s_P=C\xi^{-2}n/K_0$ this is $\tfrac{2}{C}\xi^2G^2$, so Chebyshev gives deviation $\le\tfrac{C_*}{12}
\xi G$ with probability $\ge 1-\tfrac1{100}$ for $C$ large. Hence $\abs{\widehat A_P-A_L(P)}\le\tfrac{C_*}
4\xi G$.

\emph{(ii) Support weight, $\widehat\Wq$.} \Cref{lem:support} with relative accuracy
$\eta=\dfrac{C_*\xi}{4C_W}=\Theta(\xi)$ gives $\abs{\widehat\Wq-\Wq}\le\eta\Wq\le\eta C_W G=\tfrac{C_*}4\xi
G$.

\emph{(iii) Bulk of $Q$, $\widehat A_Q$.} The support size $K=\abs V$ is not known exactly, so the estimator
scales the sample mean $\bar X=\tfrac1{s_Q}\sum_i X_L(v_i)$ by the estimate $\widehat K=(1\pm O(\xi))K$ of
\cref{sec:support-oracle}: $\widehat A_Q=\widehat K\,\bar X$. Writing $\mu=\E X_L=A_L(H_Q)/K$,
\[
  \abs{\widehat K\,\bar X-A_L(H_Q)}\le\Bigl|\tfrac{\widehat K}{K}-1\Bigr|A_L(H_Q)+\widehat K\,\abs{\bar X-\mu}
   =O(\xi)\,A_L(H_Q)+\widehat K\,\abs{\bar X-\mu},
\]
and $A_L(H_Q)\le\wt(\MST(H_Q))\le(1+\rho)\Wq=O(G)$, so the first term is $O(\xi G)$. For the second, with
$K=\Theta_\xi(K_0)$ ($K\le C_\xi K_0=O(K_0/\xi)$ by \cref{lem:kwindow}), $\Var(K\bar X)=K^2\Var(\bar X)\le
K L\,\wt(\MST(H_Q))/s_Q=O(\tfrac{K}{K_0}G^2/s_Q)=O(\xi^{-1}G^2/s_Q)$ by \eqref{eq:dt-var}, so $s_Q=
O(\xi^{-3})$ samples give $\widehat K\abs{\bar X-\mu}\le\tfrac{C_*}{12}\xi G$ with probability
$\ge1-\tfrac1{100}$ (using $\widehat K=O(K)$). So far this bounds $\abs{\widehat A_Q-A_L(H_Q)}$, the error
against the \emph{spanner}'s clipped weight; the spanner bias adds $\abs{A_L(H_Q)-A_L(Q)}\le\rho\,A_L(Q)\le
\rho\Wq=O(\xi G)\le\tfrac{C_*}{12}\xi G$ by \eqref{eq:spanner-stretch}. Adding the support-size, sampling,
and spanner-bias terms, $\abs{\widehat A_Q-A_L(Q)}\le\tfrac{C_*}4\xi G$. The extra $K/K_0=O(1/\xi)$ factor,
from the window being $\Theta_\xi$ and not $\Theta$, is what raises $s_Q$ from $O(\xi^{-2})$ to
$O(\xi^{-3})$.

\emph{(iv) Tail transfer.} $\abs{B_L(Q)-B_L(P)}\le\tfrac{C_*}4\xi G$ is \eqref{eq:tail-transfer}, which
gave $O(\xi W)$.

Summing the four through \eqref{eq:decomp} (note $W=A_L(P)+B_L(P)$ and $B_L(Q)=\Wq-A_L(Q)$),
\[
  \abs{\widehat W(G)-W}=\bigl|(\widehat A_P-A_L(P))+(\widehat\Wq-\Wq)-(\widehat A_Q-A_L(Q))
    +(B_L(P)-B_L(Q))\bigr|\le C_*\xi G,
\]
which is \eqref{eq:guess-accuracy}. Each piece fails with probability $\le\tfrac1{50}$, so one call is
within $C_*\xi G$ with probability $\ge\tfrac{9}{10}$; the median of $O(\log n)$ calls (\cref{sec:main-combine})
then makes every guess with $G\ge W$ correct simultaneously with constant probability, an aborted call
counting as a failed one.
\end{proof}

The cost of one per-guess call is $\Otil_\xi(n^{1/3})$ (the maximum of the three pieces in
\cref{sec:main-pieces}), and the $O(\log n)$-guess search multiplies this by $O(\log n)$, so
\cref{thm:main} runs in $\Otil_\eps(n^{1/3})$ range-counting queries. \qed

\subsection{Tracking the constants}\label{sec:app-constants}

For completeness we trace the dependence on the accuracy parameters through the whole pipeline, so that
the $\Otil_\eps(\cdot)$ notation can be unfolded. The chain of parameters (\cref{tab:params}) is
\[
  \eps\;\xrightarrow{\;/32\;}\;\zeta\;\xrightarrow{\;/C_*\;}\;\xi\;\xrightarrow{\;\Theta\;}\;
  \rho,\,\eta\;\xrightarrow{\;\Theta\;}\;\beta\;\xrightarrow{\;/32\;}\;\sigma,
\]
each a fixed constant multiple of its predecessor, so all are $\Theta(\eps)$ and $1/\sigma,1/\beta,1/\eta,
1/\rho,1/\xi,1/\zeta=\Theta(1/\eps)$. Write $\Lambda=\log(n\Delta/(h\eta))=\Theta(\log n)$ for the
amplification factor (using $\Delta=O(n)$ and $h\ge 1$).

\emph{Support estimator (\cref{lem:support}).} One additive call makes $\Otil_\beta(\sqrt K)$ queries; the
proof of \cref{sec:app-accounting} exhibited the explicit form $O(\beta^{-8}\sqrt K\,\Lambda^{O(1)})$,
coming from $\sigma=\beta/32$ (a $\beta^{-2}$ from the scale-graph degree), $T=O(\beta^{-1}\log(\Delta/h))$
scales, the two $T^2/\beta^2$ prefactors of \eqref{eq:cand-per-scale} and \eqref{eq:expl-per-scale}, and
the median amplification $\Lambda$. The guess-removal multiplies by the $O(\log(\Delta/h))$ guesses, and
adding the inverse-Bernoulli cost gives $\Otil_\eta(\sqrt K+n/K)=O\!\bigl(\eta^{-O(1)}(\sqrt K+n/K)
\Lambda^{O(1)}\bigr)$.

\emph{Death-time samplers.} The bulk sampler uses $s_P=C\xi^{-2}n/K_0$ seeds, each search $O(\rho^{-2}
\log^2\Delta\cdot\log n)=O(\xi^{-2}\poly\log n)$ queries by \cref{lem:incident} and the harmonic extraction
bound, for $O(\xi^{-4}(n/K_0)\poly\log n)$ queries; the support sampler uses $s_Q=O(\xi^{-3})$ seeds (the
$\xi^{-3}$, not $\xi^{-2}$, accounts for $K=\Theta_\xi(K_0)$) at $\Otil(n/K)$ each. Both are
$O(\eps^{-O(1)}n^{1/3}\poly\log n)$.

\emph{Total.} With $K=\Theta_\xi(K_0)=\Theta(n^{2/3})$ the four parts of a single call -- the
regularization search and $K$-estimation, the support estimator, and the two death-time samplers --
each cost $\Otil_\eps(n^{1/3})$ as bounded above and elsewhere in this appendix, so one global call has
\[
  \E[\text{queries per call}]=\E[\text{regularization}]+\E[\text{support estimator}]
   +\E[\text{two death-time samplers}]=O\!\bigl(\eps^{-O(1)}\,n^{1/3}\,\poly\log n\bigr),
\]
and this is exactly the quantity the deterministic global budget $B$ of \cref{sec:wspd-cap} caps. The
$O(\log n)$ guesses and the final median keep the overall bound at the same $\Otil_\eps(n^{1/3})$.
The exponent of $1/\eps$ is a fixed constant; we have not optimized it, as the contribution of the paper is
the $n^{1/3}$ rate. For fixed $\eps$ the bound is $\Otil(n^{1/3})$, and the lower bound of \cref{cor:tight}
shows the $n^{1/3}$ cannot be reduced.
