\section{The clipped death-time primitive and the reduction for \texorpdfstring{$P$}{P}}\label{sec:dt}

\Cref{lem:support} estimates the MST weight of a grid support. To estimate $W=\wt(\MST(P))$ itself we
split the MST weight at a clip level $L$ into a \emph{bulk} part (edges up to $L$) and a \emph{tail} part
(the excess of long edges), estimate the bulk directly on $P$ by a death-time sampler, and route the tail
through a snapped support handled by \cref{lem:support}. This section states the two ingredients; the
proofs are in \cref{sec:app-assembly}.

\subsection{The clipped death-time sampler}\label{sec:dt-sampler}

The support estimator integrates the cluster-count curve directly. For $P$ itself we instead sample the
MST edge weights through a device that turns ``how heavy is the MST'' into ``how far must a random vertex
travel to reach someone more senior,'' which a spanner lets us answer with a local search. The staircase
picture is \cref{fig:deathtime}: $c_S(t)-1$ counts the MST edges heavier than $t$, its area is the MST
weight, and the death times read that staircase off one vertex at a time.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.95, every node/.style={font=\footnotesize}]
  \draw[->] (0,0) -- (6.3,0) node[right] {$t$};
  \draw[->] (0,0) -- (0,3.2) node[above] {$c_S(t)-1$};
  % staircase, decreasing
  \draw[blue!60!black, very thick]
    (0,2.6) -- (1.2,2.6) -- (1.2,2.0) -- (2.1,2.0) -- (2.1,1.4) -- (3.3,1.4)
    -- (3.3,0.8) -- (4.4,0.8) -- (4.4,0.4) -- (5.4,0.4) -- (5.4,0) -- (6.0,0);
  % shaded area = MST weight
  \begin{scope}
    \clip (0,0) -- (0,2.6) -- (1.2,2.6) -- (1.2,2.0) -- (2.1,2.0) -- (2.1,1.4) -- (3.3,1.4)
      -- (3.3,0.8) -- (4.4,0.8) -- (4.4,0.4) -- (5.4,0.4) -- (5.4,0) -- cycle;
    \fill[blue!8] (0,0) rectangle (6,3);
  \end{scope}
  % clip level L
  \draw[red, dashed] (3.3,0) -- (3.3,1.7); \node[red] at (3.3,1.95) {$L$};
  \node[blue!50!black] at (1.9,2.85) {area $=\wt(\MST(S))$};
  \node at (1.5,0.9) {$A_L$}; \node at (4.6,0.55) {$B_L$};
  \draw[<->] (0,-0.35) -- (3.3,-0.35); \node at (1.65,-0.62) {bulk (clipped)};
  \draw[<->] (3.3,-0.35) -- (5.4,-0.35); \node at (4.35,-0.62) {tail};
\end{tikzpicture}
\caption{The cluster-count staircase. Each downward step is one MST edge, of width equal to its length, so
the shaded area is $\wt(\MST(S))=\int_0^\infty(c_S(t)-1)\,dt$. The clip at $L$ splits it into the bulk
$A_L=\sum_e\min\{\abs e,L\}$ (estimated on $P$ by the death-time sampler) and the tail $B_L=\sum_e(\abs
e-L)_+$ (routed through a snapped support).}
\label{fig:deathtime}
\end{figure}

Let $H$ be a connected weighted graph on $V$ with a fixed root $r$. Assign independent uniform ranks to
$V\setminus\{r\}$ and define the \emph{death time} $\tau(v)$ as the bottleneck (min--max-edge) distance
from $v$ to the root or to any lower-ranked vertex. The death times recover the MST edge weights as a
multiset:
\begin{equation}\label{eq:death-multiset}
  \set{\tau(v):v\in V}=\set 0\cup\set{\wt(e):e\in\MST(H)},
\end{equation}
because for every $t$ the count $\abs{\{v:\tau(v)>t\}}$ equals $c_H(t)-1$, the number of MST edges heavier
than $t$. A bottleneck Dijkstra search from $v$ that halts at the first lower-ranked vertex computes
$\tau(v)$ while extracting $O(\log\abs V)$ vertices in expectation. For a clip level $L$ set
$X_L(v)=\min\{\tau(v),L\}$. Then $\abs V\cdot\E X_L=A_L(H)\eqdef\sum_{e\in\MST(H)}\min\{\wt(e),L\}$, and
$X_L^2\le L X_L$, so the average of $s$ independent samples has
\begin{equation}\label{eq:dt-var}
  \Var\Bigl[\tfrac{\abs V}{s}\textstyle\sum_i X_L(v_i)\Bigr]\le\frac{\abs V\,L\,\wt(\MST(H))}{s}.
\end{equation}
We apply this to a locally navigable Euclidean $(1+\rho)$-spanner $H_X$ of a point set $X$, realized under
the oracle by a WSPD (\cref{sec:wspd}). Since $H_X$ is a Euclidean subgraph and a $(1+\rho)$-spanner,
$c_X(t)\le c_{H_X}(t)\le c_X(t/(1+\rho))$, so integrating,
\begin{equation}\label{eq:spanner-stretch}
  A_L(X)\le A_L(H_X)\le(1+\rho)\,A_L(X),
\end{equation}
i.e.\ the spanner inflates the clipped weight by at most $\rho\,\wt(\MST(X))$.

\subsection{Support regularization}\label{sec:dt-regularize}

The death-time sampler handles the bulk of $P$, but its variance \eqref{eq:dt-var} grows with the clip
level, so it is efficient only up to $L$; the long edges beyond $L$ are rare and heavy, exactly the regime
the support estimator was built for. To hand them over we need a support whose size is the balance point
$K_0=\Theta(n^{2/3})$ and whose snapping error is controlled. Choosing the grid scale is the one place a
search is needed, because the right scale depends on how the (unseen) points are spread.

Fix a guess $G\ge W$, put $K_0=\ceil{n^{2/3}}$ and $L=G/K_0$. We choose a grid scale $h$ whose support
$Q$ has $K=\Theta(K_0)$ cells. The number $K_h$ of nonempty $h$-cells obeys a packing bound analogous to
\cref{lem:packing}:
\begin{equation}\label{eq:Kh-packing}
  K_h\le\frac{8W}{h}+4 .
\end{equation}
Starting from a dyadic quadtree scale $h_0=\Theta(\xi L)$ (the largest dyadic side below $\xi L$, so $h_0$
is within a factor $2$ of $\xi L$) and halving until $K_h$ first reaches a constant multiple of $K_0$
(which is certified by a fixed batch of rejection trials, since halving $h$ at most quadruples $K_h$, and
halving keeps every scale dyadic) yields a scale with
\begin{equation}\label{eq:K-window}
  cK_0\le K\le C_\xi K_0 ,
\end{equation}
in $\Otil_\xi(n/K_0)=\Otil_\xi(n^{1/3})$ queries; \cref{lem:kwindow} proves both the window and the cost.
Fix this $h$ and its support $Q$.

\subsection{Snapping}\label{sec:dt-snap}

Snapping each point of $P$ to its $h$-cell center moves it by at most $\sqrt2\,h$, so any pairwise
distance changes by at most $\delta_s=\sqrt2\,h$, and the threshold graphs interleave:
\begin{equation}\label{eq:snap-interleave}
  c_P(t+\delta_s)\le\cQ(t)\le c_P(t-\delta_s)\qquad(t\ge\delta_s).
\end{equation}
Writing $B_L(X)=\int_L^\infty(c_X(t)-1)\,dt=\sum_{e\in\MST(X)}(\abs e-L)_+$ for the tail weight,
\eqref{eq:snap-interleave} together with $c_P(t)-1\le W/t$ gives, for $\delta_s\le L/2$,
\begin{equation}\label{eq:tail-transfer}
  \abs{B_L(Q)-B_L(P)}\le\int_{L-\delta_s}^{L+\delta_s}\frac Wt\,dt\le\frac{3\delta_s}{L}W=O(\xi W),
\end{equation}
by $\delta_s=\sqrt2\,h=O(\xi L)$ (\cref{lem:kwindow} gives $h\le h_0=\Theta(\xi L)$). Finally, contracting $\MST(P)$ onto the $h$-cells and using \eqref{eq:Kh-packing}
with $K=\Omega(K_0)$ gives $\Wq\le W+\delta_s(K-1)\le C_W W$ for an absolute constant $C_W$.

\subsection{The decomposition}\label{sec:dt-decomp}

We can now state precisely which three quantities the algorithm estimates and how they combine. The
guiding identity is that the MST weight splits as bulk plus tail at the clip level, and that the tail of
$P$ equals the tail of the snapped support up to the transfer error \eqref{eq:tail-transfer}; the support
tail in turn is its total weight minus its own bulk, both of which we can estimate.

Write $A_L(X)=\sum_{e\in\MST(X)}\min\{\abs e,L\}$, so $\wt(\MST(X))=A_L(X)+B_L(X)$. We use the bulk of $P$
and the tail of the snapped support:
\begin{equation}\label{eq:decomp}
  W=A_L(P)+B_L(P),\qquad B_L(Q)=\Wq-A_L(Q),
\end{equation}
and transfer the tail by \eqref{eq:tail-transfer}. The three estimable pieces are $A_L(P)$ (death-time
sampler on $P$), $\Wq$ (\cref{lem:support}), and $A_L(Q)$ (death-time sampler on $Q$); we assemble them in
\cref{sec:main}.
