\section{Introduction}\label{sec:intro}

The weight of a Euclidean minimum spanning tree is one of the oldest statistics of a point set, and
estimating it without reading the whole input is a recurring sublinear-algorithms question. For $n$ points
in the plane the tree itself has $n-1$ edges, but its total length is a single number, and a long line of
work asks how cheaply that number can be approximated. The answer depends entirely on how the algorithm is
allowed to touch the points. Chazelle, Rubinfeld, and Trevisan~\cite{CRT05} showed that for a graph given
by adjacency lists the MST weight is $(1\pm\eps)$-estimable in time independent of the number of vertices,
and Czumaj, Sohler, and coauthors~\cite{Czumaj05,CzumajSohler09} carried the program into geometry, where
the cost is measured in queries to a geometric oracle.

We work in the \emph{orthogonal range-counting} model studied by Driemel, Monemizadeh, Oh, Staals, and
Woodruff~\cite{DMOSW25}: the point set $P\subseteq[\Delta]^2$ (with $\Delta=O(n)$) is accessed only
through a data structure that, given an axis-aligned rectangle $R$, returns $\abs{P\cap R}$, and the cost
of an algorithm is the number of such queries. This is the native primitive of a spatial database, and it
is strictly weaker than the distance and nearest-neighbor oracles used in earlier geometric MST work: a
range count reveals how many points fall in a box but neither which points nor where. Driemel et
al.~\cite{DMOSW25} gave an $\Otil(\sqrt n)$-query algorithm in this model together with an
$\Omega(n^{1/3})$-query lower bound, and left the polynomial gap between the two as the open question. We
close it.

\subsection{Our results}\label{sec:intro-results}

Our main result is a range-counting algorithm whose query cost matches the lower bound up to
polylogarithmic factors.

\begin{theorem*}[\cref{thm:main}]
For every fixed $\eps\in(0,1)$ and every $P\subseteq[\Delta]^2$ with $\abs P=n$ and $\Delta=O(n)$, there
is a randomized algorithm that, with probability at least $2/3$, returns $\widehat W$ with
$(1-\eps)\,\wt(\MST(P))\le\widehat W\le(1+\eps)\,\wt(\MST(P))$ using $\Otil_\eps(n^{1/3})$ orthogonal
range-counting queries.
\end{theorem*}

The success probability is boosted to $1-\delta$ by $O(\log(1/\delta))$ independent repetitions and a
median. Since Driemel et al.~\cite{DMOSW25} prove that $\Omega(n^{1/3})$ queries are necessary even to
estimate $\wt(\MST(P))$ within a constant factor, \cref{thm:main} settles the range-counting query
complexity of Euclidean MST-weight estimation up to polylogarithmic and $\eps$-dependent factors
(\cref{sec:optimality}). \Cref{tab:prior} places the bound in context.

\begin{table}[t]
\centering
\small
\begin{tabular}{llll}
\toprule
result & oracle & queries & error\\
\midrule
Czumaj et al.~\cite{Czumaj05} & range-emptiness $+$ cone-ANN & $\Otil(\sqrt n)$ & $O(n^{1/4})$ rel.\\
Driemel et al.~\cite{DMOSW25} & orthogonal range counting & $\Otil(\sqrt n)$ & $(1\pm\eps)$\\
Driemel et al.~\cite{DMOSW25} (LB) & orthogonal range counting & $\Omega(n^{1/3})$ & const.\ mult.\\
\textbf{this paper} & orthogonal range counting & $\boldsymbol{\Otil_\eps(n^{1/3})}$ & $(1\pm\eps)$\\
\bottomrule
\end{tabular}
\caption{Sublinear estimation of planar Euclidean MST weight under geometric oracles.}
\label{tab:prior}
\end{table}

The engine behind \cref{thm:main} is a self-contained estimator for the MST weight of a grid
\emph{support}, which we isolate because it is where the improvement lives.

\begin{lemma*}[\cref{lem:support}]
For every $\eta\in(0,\tfrac14)$ there is a randomized algorithm that, with probability at least $2/3$,
estimates the MST weight $\Wq$ of the set $Q$ of nonempty-cell centers of a grid imposed on $P$ to within
a $(1\pm\eta)$ factor, using $\Otil_\eta(\sqrt K+n/K)$ range-counting queries, where $K=\abs Q$.
\end{lemma*}

Balancing the two terms of \cref{lem:support} at $K=\Theta(n^{2/3})$ gives the $n^{1/3}$ rate; the
reduction from $P$ to a support of this size, and the routing of the heavy tail of long MST edges, is the
content of \cref{sec:dt,sec:main}.

\subsection{Techniques}\label{sec:intro-tech}

\paragraph{The barrier, and sampling space instead of mass.}
By the Chazelle--Rubinfeld--Trevisan identity~\cite{CRT05}, the MST weight is the area under the
cluster-count curve, $\wt(\MST(S))=\int_0^\infty(c_S(t)-1)\,dt$, where $c_S(t)$ counts the single-linkage
clusters of $S$ at scale $t$ (\cref{sec:prelim}). Estimating the curve by sampling points and exploring
their neighborhoods is the natural approach, and it is exactly what costs $\Omega(\sqrt n)$: a few far-flung
points can carry a constant fraction of the weight while occupying a vanishing fraction of the mass, so a
mass sample misses them with high probability and undercounts. Our estimator samples \emph{space} rather
than mass. At each scale we impose a fine grid and draw candidate cells uniformly from a spatial universe
that \emph{includes empty cells}; emptiness is one cheap range count, and on an occupied cell we run a
min-rank exploration that returns a nonzero value only at the unique \emph{leader} of its component. A
heavy outlier occupies its own cell and is therefore sampled with the right probability, not with
probability proportional to its (tiny) mass. This \emph{empty-cell spatial leader estimator}
(\cref{sec:support-estimator}) is the one new primitive of the paper: it has expectation equal to the
component count and variance bounded by $M\cdot c$, with $M$ the universe size and $c$ the component count,
so the trial count is governed by the number of components and not by the number of occupied cells. An
active-cover packing inequality, $\Wq\ge b\delta/16$ (\cref{lem:packing}), then collapses the geometric
sum of per-scale costs to $\Otil(\sqrt K)$.

\paragraph{From a support back to the point set.}
\Cref{lem:support} estimates a grid support, not $P$. To recover $W=\wt(\MST(P))$ we split the MST weight
at a clip level $L$ into a bulk part (edges up to $L$) and a tail part (the excess of long edges). The
bulk is estimated directly on $P$ by a \emph{clipped death-time sampler}: assigning random ranks to
vertices and reading off, for a sampled vertex, the bottleneck distance to the nearest lower-ranked
vertex recovers the MST edge weights as a multiset (\cref{sec:dt-sampler}), and clipping at $L$ controls
the variance. The tail is routed through a grid support obtained by \emph{regularization and snapping}
(\cref{sec:dt-regularize,sec:dt-snap}): we choose a grid scale whose support has $\Theta(n^{2/3})$ cells,
snap $P$ to cell centers, and bound the weight distortion by a packing argument and an interleaving of
threshold graphs. The three estimable pieces, the bulk of $P$, the support weight $\Wq$, and the bulk of
the support, assemble into $\widehat W=\widehat A_P+\widehat\Wq-\widehat A_Q$ with a controlled error
budget and a geometric search for the unknown scale of $W$ (\cref{sec:main}). The death-time sampler runs
on a locally navigable $(1+\rho)$-spanner realized implicitly under the oracle by a well-separated pair
decomposition; we give the full range-counting implementation, including the neighbor oracle and the
bottleneck search, in \cref{sec:wspd}.

\subsection{Related work}\label{sec:intro-related}

\paragraph{Sublinear MST weight.}
The graph model was opened by Chazelle, Rubinfeld, and Trevisan~\cite{CRT05}, who reduced MST weight to
counting connected components of threshold graphs and gave a $(1\pm\eps)$-estimator in
$\Otil(d_{\max}\eps^{-2})$ time for average degree $d_{\max}$; Czumaj and Sohler~\cite{CzumajSohler09}
extended the component-counting reduction to general metrics under a distance oracle, and Patlin and van
den Brand~\cite{PatlinvdB25} recently revisited the running-time dependence. The cluster-count integral
and the threshold-graph view we use throughout descend from this line.

\paragraph{Euclidean MST estimation and geometric oracles.}
Czumaj, Erg\"un, Fortnow, Magen, Newman, Rubinfeld, and Sohler~\cite{Czumaj05} gave the first sublinear
planar Euclidean MST-weight estimator, in $\Otil(\sqrt n)$ time, using range-emptiness queries together
with approximate cone nearest-neighbor search and obtaining a polynomial additive error; the same group's
metric reduction~\cite{CzumajSohler04,CzumajSohler09} and the early sublinear geometric algorithms of
Chazelle, Liu, and Magen~\cite{ChazelleLiuMagen03} set the stage. Arya, Mount, and
Park~\cite{AryaMountPark15} studied geometric MST range queries with approximate-range data structures,
building on approximate range searching~\cite{AryaMount00}, and range-clustering and approximate-range
clustering queries were studied by Abrahamsen et al.~\cite{AbrahamsenEtal17} and Oh and
Ahn~\cite{OhAhn18}. The range-counting oracle of Driemel et al.~\cite{DMOSW25} is the setting of the
present paper; their $\Otil(\sqrt n)$ upper bound also reduces to component counting, via the
Callahan--Kosaraju WSPD spanner~\cite{CK95,HarPeled11,NarasimhanSmid07}, but pays $\sqrt n$ at the
component-counting step, which is the cost our spatial leader estimator removes. The oracle has since
been applied to convex-hull and emptiness problems~\cite{SchiblerXueZhu26}. Orthogonal range counting is
itself a heavily studied data-structuring primitive~\cite{AfshaniArgeLarsen09,ChanLarsenPatrascu11,
ChanTsakalidis18,Chan19,ShengTao11,deBergEtal08}.

\paragraph{Single-linkage clustering and related primitives.}
The components of the threshold graph are the single-linkage clusters, and the cost of single-linkage
clustering has been studied directly in the sublinear regime by Peng, Sohler, and
Xu~\cite{PSX25}; MST and clustering under strong and weak distance oracles were studied
in~\cite{MetricOracleMST23}. Streaming Euclidean MST to a constant factor was given by Chen, Cohen-Addad,
Jayaram, Levi, and Waingarten~\cite{ChenCAJLW23}; the contraction trick that keeps component counts
robust to additive error, which we reuse in the support regularization, traces to dynamic-stream sampling
of Frahling, Indyk, and Sohler~\cite{FrahlingIndykSohler05} and Indyk~\cite{Indyk04} and to
relative-error $L_p$-sampling~\cite{MonemizadehWoodruff10}. Component counting and related sublinear
estimators are part of the broader property-testing toolkit~\cite{CzumajSohler01,CzumajSohler10,
EdenRonSeshadhri18,OnakEtal12,RubinfeldShapira11,BehnezhadRoghaniRubinstein23,CormodeEtal17}, with
concentration via Chernoff bounds~\cite{Chernoff52}. The well-separated pair decomposition and geometric spanners are standard tools
(Callahan--Kosaraju~\cite{CK95}, Har-Peled~\cite{HarPeled11}, Narasimhan--Smid~\cite{NarasimhanSmid07}),
and telescoping range-count sampling is due to Monemizadeh~\cite{Monemizadeh23}.

The paper is organized as the techniques above suggest: \cref{sec:prelim} fixes the model and the
cluster-count integral; \cref{sec:support} states and uses the support-MST estimator; \cref{sec:dt}
gives the death-time primitive and the reduction for $P$; \cref{sec:main} assembles the main theorem;
\cref{sec:optimality} records the match to the lower bound; and \cref{sec:open} collects open questions.
Full proofs are in \cref{sec:wspd,sec:app-lemma3,sec:app-assembly}.
