\section{Estimating the MST weight of \texorpdfstring{$P$}{P}}\label{sec:main}

We now assemble the pieces of \cref{sec:dt} into the main result.

\begin{theorem}[Main]\label{thm:main}
For every fixed $\eps\in(0,1)$ and every $P\subseteq[\Delta]^2$ with $\abs P=n$ and $\Delta=O(n)$, there
is a randomized algorithm that, with probability at least $2/3$, outputs $\widehat W$ with
\[
  (1-\eps)\,\wt(\MST(P))\ \le\ \widehat W\ \le\ (1+\eps)\,\wt(\MST(P)),
\]
using $\Otil_\eps\!\bigl(n^{1/3}\bigr)$ range-counting queries to $P$.
\end{theorem}

\begin{algorithm}[t]
\caption{Estimating $W=\wt(\MST(P))$ to a $(1\pm\eps)$ factor}\label{alg:main}
\KwIn{oracle access to $P\subseteq[\Delta]^2$; accuracy $\eps$}
\KwOut{$\widehat W$ with $(1-\eps)W\le\widehat W\le(1+\eps)W$ w.p.\ $\ge 2/3$}
$U\leftarrow\sqrt2(n-1)\Delta$; $\;\zeta_0\leftarrow\tfrac1{64}$\;
\ForEach{$G=U,U/2,U/4,\dots$ \textnormal{(at most $O(\log n)$ guesses)}}{
  $\widehat W(G)\leftarrow$ median of $O(\log n)$ runs of \textsc{Estimate}$(G,\zeta_0)$ \tcp*{aborted run $=+\infty$}
  \lIf{$\widehat W(G)\ge G/8$}{$G_0\leftarrow G$; \textbf{break}}
}
\Return median of $O(\log n)$ runs of \textsc{Estimate}$(G_0,\eps/32)$\;
\medskip
\textbf{procedure} \textsc{Estimate}$(G,\zeta)$, with $\xi=\zeta/C_*$, $K_0=\ceil{n^{2/3}}$, $L=G/K_0$:\;
\Indp
regularize to a support $Q$ with $K=\Theta_\xi(K_0)$ cells (\cref{lem:kwindow}); estimate $K$\;
$\widehat A_P\leftarrow$ clipped death-time sampler on a $(1{+}\rho)$-spanner of $P$, $s_P=O(\xi^{-2}n^{1/3})$ seeds\;
$\widehat\Wq\leftarrow$ support-MST estimator (\cref{lem:support}) at relative accuracy $\Theta(\xi)$\;
$\widehat A_Q\leftarrow$ clipped death-time sampler on a $(1{+}\rho)$-spanner of $Q$, $s_Q=O_\xi(1)$ seeds\;
\Return $\widehat A_P+\widehat\Wq-\widehat A_Q$ (or \textsc{fail} if the global query budget is exceeded)\;
\end{algorithm}

\paragraph{Probability and budget, in one place.}
Three notions of cost appear in the analysis and we state how they fit together. The estimators are
randomized and their natural query counts are bounds in expectation or with high probability; the algorithm
converts these into a \emph{deterministic} per-run bound by carrying a global query budget $B$ and aborting
once it is spent (\cref{sec:wspd-cap}), and an aborted run is counted as a failed sample inside the median
amplification. The following lemma packages the guarantee of one call; \cref{thm:main} then follows by the
geometric search over guesses.

\begin{lemma}[Per-call guarantee]\label{lem:percall}
Fix a guess $G$ and set $\zeta=C_*\xi$. One run of $\textsc{Estimate}(G,\zeta)$ makes at most a
\emph{deterministic} $B=\Otil_\zeta(n^{1/3})$ range counts, and its expected count is also
$\Otil_\zeta(n^{1/3})$. If $G\ge W$ then with probability at least $9/10$ the run completes within $B$ and
returns $\widehat W(G)$ with $\abs{\widehat W(G)-W}\le\zeta G$; the complementary \emph{bad} event, of
probability at most $1/10$, is that the run either overruns $B$ (returning \textsc{fail}) or completes with
an inaccurate value.
\end{lemma}

\begin{proof}
A run estimates the three pieces of \cref{sec:main-pieces}, each to additive error $O(\xi G)$ on the
$\poly\log n$ internal estimates (the $T=\Otil(1)$ scales of the support call and the death-time means),
which we amplify to failure probability $o(1/\log n)$ each by a median of $O(\log n)$ blocks and union-
bound; this gives $\abs{\widehat W(G)-W}\le C_*\xi G=\zeta G$ with probability $\ge 1-\tfrac1{20}$ whenever
$G\ge W$ (\cref{sec:app-combine}, \eqref{eq:guess-accuracy}). The total expected query count of the four
parts -- regularization and $K$-estimation, the support estimator, and the two death-time samplers -- is
$\Otil_\zeta(n^{1/3})$ (\cref{sec:app-constants}). Setting the deterministic budget $B$ to a large constant
times this expectation and aborting on overrun makes every run cost at most $B$, and by Markov a run
overruns with probability $\le\tfrac1{20}$; folding the two failure modes gives the stated $9/10$ and
$1/10$.
\end{proof}

\subsection{Estimating the three pieces under a guess}\label{sec:main-pieces}

Fix a guess $G\ge W$ and an internal accuracy $\xi$, and set $K_0=\ceil{n^{2/3}}$, $L=G/K_0$, with the
regularized support $Q$ of \cref{sec:dt-regularize} ($K=\Theta_\xi(K_0)$). The decomposition
\eqref{eq:decomp} expresses $W$ through three estimable quantities, $A_L(P)$, $\Wq$, and $A_L(Q)$. We
estimate each to additive error $O(\xi G)$; the common normalization by $G$ (not by the individual
quantities) is what lets the same internal accuracy serve all three and the geometric search below. The
clip level $L=G/K_0$ is the lever: it is small enough that the death-time variance \eqref{eq:dt-var},
which scales with $L$, stays affordable, and large enough that the tail beyond $L$ carries only the heavy
edges the support handles. Each estimate:
\begin{itemize}[leftmargin=1.5em,itemsep=2pt]
\item \emph{Bulk $A_L(P)$.} Run the death-time sampler on a $(1+\rho)$-spanner of $P$ with
  $\rho=\Theta(\xi)$. By \eqref{eq:dt-var} with $\abs V=n$, $L=G/K_0$ and $\wt(\MST(P))=W\le G$, the
  variance is $O(nG^2/(K_0 s_P))$, so $s_P=O(\xi^{-2}n/K_0)=O(\xi^{-2}n^{1/3})$ samples suffice; with
  $\Otil(1)$ queries per sample this is $\Otil_\xi(n^{1/3})$ queries (the spanner bias is
  $\le\rho W=O(\xi G)$ by \eqref{eq:spanner-stretch}).
\item \emph{Support weight $\Wq$.} Apply \cref{lem:support} with relative error $\Theta(\xi/C_W)$; since
  $\Wq\le C_W W\le C_W G$ this gives additive error $O(\xi G)$ in $\Otil_\xi(\sqrt K+n/K)=\Otil_\xi(n^{1/3})$
  queries.
\item \emph{Support bulk $A_L(Q)$.} Run the death-time sampler on a $(1+\rho)$-spanner of $Q$. Here
  $\abs V=K=\Theta_\xi(K_0)$ (\cref{lem:kwindow}; $K$ may be as large as $C_\xi K_0=\Theta(K_0/\xi)$),
  $L=G/K_0$ and $\wt(\MST(Q))=\Wq=O(G)$, so \eqref{eq:dt-var} makes the variance $O(\tfrac{K}{K_0}G^2/s_Q)
  =O(\xi^{-1}G^2/s_Q)$; thus $s_Q=O(\xi^{-3})=O_\xi(1)$ support samples suffice, the extra $K/K_0=O(1/\xi)$
  factor absorbed in $O_\xi(\cdot)$. Each uniform support sample costs $\Otil(n/K)=\Otil_\xi(n^{1/3})$
  queries (\cref{sec:support-oracle}).
\end{itemize}
The assembly does not need the three errors to be independent: we only bound each by $O(\xi G)$ with
constant probability and apply a union bound, so even though the regularized support $Q$ enters both
$\widehat\Wq$ and $\widehat A_Q$, no independence is assumed.


\begin{remark}[The choice $K_0=n^{2/3}$]\label{rem:k0}
The support size $K_0$ is the only free knob, and the three costs above pin it down. The support estimator
costs $\Otil(\sqrt{K_0}+n/K_0)$; the bulk sampler costs $\Otil(\xi^{-2}n/K_0)$ (more support cells mean a
smaller clip $L=G/K_0$, hence lower death-time variance and fewer seeds); the support sampler costs
$\Otil(n/K_0)$ per seed. The first is increasing in $K_0$ through its $\sqrt{K_0}$ term and the rest are
decreasing, and all four meet when $\sqrt{K_0}=\Theta(n/K_0)$, i.e.\ $K_0=\Theta(n^{2/3})$, giving the
common value $\Otil(n^{1/3})$. Any other choice makes one of the terms dominate and the bound worse.
\end{remark}

\subsection{Combining and removing the guess}\label{sec:main-combine}

Output
\begin{equation}\label{eq:combine}
  \widehat W(G)=\widehat A_P+\widehat\Wq-\widehat A_Q .
\end{equation}
By the decomposition \eqref{eq:decomp} and the tail transfer \eqref{eq:tail-transfer},
\[
  \abs{\widehat W(G)-W}\le\abs{\widehat A_P-A_L(P)}+\abs{\widehat\Wq-\Wq}+\abs{\widehat A_Q-A_L(Q)}
    +\abs{B_L(Q)-B_L(P)},
\]
and each of the four terms is $O(\xi G)$, so for a suitable constant $C_*$,
\begin{equation}\label{eq:guess-accuracy}
  \abs{\widehat W(G)-W}\le C_*\xi G\qquad\text{whenever }G\ge W .
\end{equation}
This is \eqref{eq:guess-accuracy}, the per-run accuracy that \cref{lem:percall} packages with the
deterministic budget. We amplify each guess by taking the median $\widehat W(G)$ of $t=O(\log n)$
independent runs, an aborted run contributing $+\infty$. By \cref{lem:percall}, for $G\ge W$ a run is
\emph{bad} -- it aborts or is inaccurate -- with probability at most $\tfrac1{10}<\tfrac12$, so fewer than
half the runs are bad with probability $1-n^{-\Omega(1)}$ and the median is accurate and finite.
A union bound over the $O(\log n)$ guesses keeps all guesses with $G\ge W$ simultaneously accurate;
guesses with $G<W$ need no guarantee beyond not stopping the search early, which the accurate guesses
enforce.

The estimator needs a guess $G\ge W$ to set its clip level $L=G/K_0$ and its budgets, but $W$ is what we
are computing. The way out is that the guarantee \eqref{eq:guess-accuracy} is one-sided in a useful way:
for any $G\ge W$ the output $\widehat W(G)$ is within $\zeta G$ of $W$, so reading $\widehat W(G)$ against
the scale of $G$ tells us whether $G$ is still far above $W$ or has come down near it. Sweeping $G$
downward by halving and watching for $\widehat W(G)$ to first reach a constant fraction of $G$ therefore
brackets $W$ within a factor of two, after which one calibrated call finishes the job. We make this
precise.

A deterministic upper bound is $U=\sqrt2(n-1)\Delta\ge W$, and $W\ge n-1$ since the points are distinct
integer-grid points, so $U/W=O(\Delta)=O(n)$ and only $O(\log n)$ guesses are needed. Run the per-guess
estimator with $\zeta_0=\tfrac1{64}$ on $G=U,U/2,U/4,\dots$, amplified so that with constant probability
every call with $G\ge W$ is correct, and stop at the first $G$ with $\widehat W(G)\ge G/8$. If $G\ge16W$
then $\widehat W(G)\le W+G/64\le G/16+G/64<G/8$, so the search does not stop; if $G\in[W,2W)$ then
$\widehat W(G)\ge W-G/64>G/8$, so it stops while $G\ge W$. Hence the selected $G_0$ satisfies $W\le
G_0<16W$. A final call on $G_0$ with $\zeta=\eps/32$, amplified by the same median of $O(\log n)$ runs,
gives $\abs{\widehat W-W}\le(\eps/32)G_0<(\eps/2)W$. Allotting each of the $O(\log n)$ guess calls and the
final call a failure budget $O(1/\log n)$ keeps the whole procedure correct with probability $\ge 2/3$,
and the $O(\log n)$ guesses keep the total cost $\Otil_\eps(n^{1/3})$. This proves \cref{thm:main}; the
deferred variance, packing, and snapping calculations are completed in \cref{sec:app-assembly}.

\begin{remark}[Running time]\label{rem:time}
Although \cref{thm:main} counts oracle queries, the algorithm is also efficient in the usual sense. Every
step between queries is bookkeeping on cells and ranks: the cover construction, the offline candidate
generation, the leader explorations, the WSPD partner enumeration, and the bottleneck priority queue each
do work proportional to the number of cells touched, which is within a polylogarithmic factor of the query
count. On a word RAM with $O(\log n)$-bit words, where a quadtree cell, a representative's integer coordinates,
and a rank fit in $O(1)$ words, the total running time is therefore $\Otil_\eps(n^{1/3})$ as well,
dominated by the queries up to polylogarithmic factors. The only numeric operations are comparisons and
additions of edge lengths in the bottleneck search and the integration: edge-length comparisons are done
on \emph{squared} Euclidean distances, which are exact integers, and the clipped-weight sums need each
length only to relative precision $\poly(\eps/\log n)$, supplied by $O(\log(n/\eps))$-bit approximations of
the square roots. So no exact irrational arithmetic is required, and the algorithm runs in the standard
integer word RAM model.
\end{remark}

\subsection{The algorithm on a concrete instance}\label{sec:main-example}

To see the pieces act in concert, follow them on the hard shape of \cref{sec:overview}, with parameters
that stay inside the model $\Delta=O(n)$. The input is a unit-spaced background carpet of $n-s$ points
filling a $\Theta(\sqrt n)\times\Theta(\sqrt n)$ region, together with only $s=\Theta(\log n)$
\emph{satellite} points placed so that the MST links them by a chain of geometrically growing edges of
lengths $2^1,2^2,\dots,2^s$, the longest reaching $2^s=\Theta(\Delta)=\Theta(n)$. The carpet contributes
$\Theta(n)$ to the MST weight (each point joins a neighbor at distance $\Theta(1)$) and the chain
contributes $\Theta(\sum_{i\le s}2^i)=\Theta(2^s)=\Theta(n)$, so $W=\Theta(n)$, with a constant fraction of
it living in the few long satellite edges.

Regularization fixes $K_0=\Theta(n^{2/3})$ and a grid at which the carpet collapses to $\Theta_xi(K_0)$ cells
while each of the $s$ far-apart satellites still occupies its own. The contrast is now sharp at the
support: the satellites are a $s/K_0=\Theta(\log n/n^{2/3})$ fraction of the support cells, so a sample of
$\poly\log$ support \emph{points} misses them, yet the long satellite edges carry a constant fraction of
$\Wq$. The leader estimator samples cells instead, where a satellite cell is one candidate among $M$ and is
detected at the right rate, so \cref{lem:support} reads the full $\Wq=\Theta(n)$ -- heavy tail included --
in $\Otil(n^{1/3})$ queries. Assembling $\widehat W=\widehat A_P+\widehat\Wq-\widehat A_Q$ keeps the
satellite tail (carried by $\widehat\Wq$) and cancels the duplicated bulk, returning $\Theta(W)$. A
point-mass algorithm at the same budget never samples a satellite and reports only the carpet weight,
missing the constant-fraction tail; the spatial leader estimator is exactly what averts that failure.
