\section{Open questions}\label{sec:open}

\paragraph{Earth Mover's Distance.}
The Earth Mover's Distance (EMD) between two planar point sets sits in the same range-counting framework of
Driemel et al.~\cite{DMOSW25} as MST weight. Their EMD results give an $O(\log\Delta)$ multiplicative
approximation with an additive-error trade-off that is tight in its own regime; whether a near-$(1\pm\eps)$
estimate of planar EMD is achievable under a pure range-counting oracle, and at what query rate, is open.
Sublinear and sketching algorithms for EMD have a long
history~\cite{AndoniDoBaIndykWoodruff09,DoBaEtal11,Charikar02}, as do near-linear geometric matching
algorithms~\cite{AgarwalChangRaghvendraXiao22,AgarwalChangXiao19}, but none operates under a pure
range-counting oracle.

We believe the empty-cell spatial leader estimator of \cref{sec:support-estimator} is the right primitive
here, and sketch why. The estimator is not specific to single-linkage clustering: at its core it estimates
the number of connected components of an \emph{implicitly defined} geometric graph from a sample of cells
drawn from a spatial universe that includes empty cells, with variance governed by the component count and
not the occupied-cell count. EMD has a dyadic structure parallel to the cluster-count integral. Writing
the two point sets as red and blue with equal mass, the cost of an optimal transport admits a quadtree
approximation: at each dyadic level one pays, per cell, the absolute red-minus-blue mass imbalance times
the cell diameter, and summing over levels approximates EMD within an $O(\log\Delta)$ factor
(the source of the multiplicative $\log\Delta$ loss already present in~\cite{DMOSW25}). The analogue of
the cluster-count curve is thus the imbalance profile across scales, and the analogue of estimating
$c_S(t)$ at a scale is estimating the number of cells with nonzero imbalance, or more precisely the total
imbalance, at that scale.

Two ingredients would need EMD analogues, which we state as open problems, not claims. First, a
\emph{spatial estimator} for the per-scale \emph{absolute} imbalance $\sum_{\text{cells}}\abs{\text{red}-
\text{blue}}$ (the quantity the dyadic transport sum charges; the signed differences cancel over a level
and are not what one wants). One would draw a candidate cell uniformly and return $\abs{\text{red}-
\text{blue}}$ for that cell scaled by the universe size, an unbiased estimator of the level's absolute
imbalance whose variance is governed by a spatial, not mass-proportional, sample; whether its variance and
a packing control can be made to behave as well as the leader estimator's $Mc$ bound is the open point.
Second, a \emph{packing inequality} relating the imbalance profile across scales to the EMD itself, playing
the role of $\Wq\ge b\delta/16$ and collapsing the geometric sum of per-scale costs. We conjecture that
establishing these and balancing the resulting $\sqrt K+n/K$ trade-off would give an $\Otil(n^{1/3})$-query
estimate of the dyadic transport approximation of planar EMD, hence an $O(\log\Delta)$-approximation of EMD
at that rate; whether the spatial-sampling idea reaches the true matching cost, and what the right
range-counting lower bound for EMD is, are the deeper questions.

\paragraph{Higher dimensions.}
For $P\subseteq[\Delta]^d$ with $d\ge3$ the WSPD spanner and the cluster-count integral both go through,
but the packing exponent and the balance point $K=\Theta(n^{2/3})$ are tuned to the plane. The
dimension-dependence of the optimal range-counting exponent, and whether it degrades gracefully with $d$,
is open.

\paragraph{The cost of learning $K$.}
The support estimator pays $\Otil(\sqrt K+n/K)$, and the additive $n/K$ is purely the cost of estimating
the support size $K$ by rejection sampling, which is what forces the balance at $K=\Theta(n^{2/3})$. If $K$
were known, or estimable more cheaply, the support estimator would cost $\Otil(\sqrt K)$ alone, and the
reduction could in principle use a smaller support. Whether the $n/K$ term is necessary, or whether the
support size can be learned in $o(n/K)$ queries (perhaps amortized across the geometric search, which
already touches the support at many scales), is open, and a positive answer might lower the exponent below
$1/3$ for inputs whose support is easy to count, though never below the worst-case $\Omega(n^{1/3})$.

\paragraph{Polylogarithmic factors and determinism.}
\Cref{cor:tight} leaves a polylogarithmic and $\eps$-dependent gap. Pinning down the exact dependence on
$\eps$ and $\log\Delta$, and asking whether a deterministic range-counting algorithm can match the
$n^{1/3}$ rate or whether randomization is necessary at this oracle, are concrete questions about the
constant in the exponent's neighborhood. The empty-cell leader estimator is fundamentally randomized (it
relies on uniform ranks and spatial sampling), so a deterministic $n^{1/3}$ algorithm, if one exists,
would likely need a different idea for the component-counting step.
