\section{The support-MST estimator}\label{sec:support}

The technical core of the algorithm estimates the MST weight of an implicitly defined grid \emph{support}.
Fix an aligned grid $\mathcal G_h$ of side $h$ on $[\Delta]^2$ (cells half-open) and let
\[
  Q=\set{z_C : C\in\mathcal G_h,\ P\cap C\ne\varnothing},\qquad K=\abs Q,
\]
where $z_C$ is the center of cell $C$. The support $Q$ is never written down: the oracle sees $P$, and a
rectangle is nonempty in $Q$ exactly when it is nonempty in $P$ (\cref{sec:support-oracle}). We estimate
$\Wq=\wt(\MST(Q))$ in a number of queries sublinear in $K$.

\begin{lemma}[Support-MST estimator]\label{lem:support}
For every $\eta\in(0,\tfrac14)$ there is a randomized algorithm that, with probability at least $2/3$,
outputs $\widehat\Wq$ with $\abs{\widehat\Wq-\Wq}\le\eta\,\Wq$ using
$\Otil_\eta\!\bigl(\sqrt K+\tfrac nK\bigr)$ range-counting queries to $P$. If $K$ is supplied to relative
error $O(\eta)$, the $n/K$ term is unnecessary.
\end{lemma}

We prove \cref{lem:support} in this section, deferring the full query accounting and the accuracy
calculation to \cref{sec:app-lemma3}.

\subsection{Oracle operations on the support}\label{sec:support-oracle}

The support $Q$ is implicit, but three operations on it reduce to range counts on $P$.

\emph{Emptiness.} Every cell and rectangle used below is a union of $h$-cells, and such a region $R$
contains a support center iff it contains a point of $P$ (the center of an $h$-cell is in $Q$ exactly when
that $h$-cell is nonempty in $P$). So $Q\cap R\ne\varnothing$ iff $P\cap R\ne\varnothing$, and one range
count on $P$ tests support emptiness.

\emph{Uniform support cell.} A uniform support cell is drawn by rejection. Sample a uniform point $p$ of
$P$ with $O(\log\Delta)$ counts (binary search on nested rectangles, \cref{sec:prelim-sample}), let $C(p)$
be its $h$-cell, and accept with probability $1/\abs{P\cap C(p)}$ (the count $\abs{P\cap C(p)}$ is one more
range query). Conditioned on acceptance, a fixed nonempty $h$-cell $C$ is returned with probability
proportional to $\sum_{p\in P\cap C}\Pr[p\text{ sampled}]\cdot\tfrac1{\abs{P\cap C}}=\abs{P\cap C}\cdot
\tfrac1n\cdot\tfrac1{\abs{P\cap C}}=\tfrac1n$, the same for every nonempty cell, so the accepted cell is
uniform over the $K$ support cells. The acceptance probability is $\sum_{C}\abs{P\cap C}\cdot\tfrac1n\cdot
\tfrac1{\abs{P\cap C}}=K/n$.

\emph{Estimating $K$.} Each rejection trial accepts independently with probability $K/n$, so $K$ is
recovered by \emph{inverse-binomial} sampling, which needs no a-priori knowledge of $K$. Fix a target
$a=\Theta(\kappa^{-2}\log(1/\delta))$ and draw trials until $a$ of them have accepted; if $T$ is the number
of trials used, output $\widehat K=a\,n/T$. The count $T$ is a negative-binomial variable with mean
$a\,n/K$, and by its concentration $T\in(1\pm\kappa)a\,n/K$ with probability $1-\delta$, so $\widehat K\in
(1\pm O(\kappa))K$ with probability $1-\delta$; on that event the procedure made $T=O(\kappa^{-2}(n/K)
\log(1/\delta))$ trials, an \emph{$\Otil(n/K)$ query bound that holds with high probability}, not only in
expectation. We run this under the same global budget as the rest of the call (\cref{sec:support-call}): if
$T$ exceeds $C\cdot\kappa^{-2}(n/K_{\min})\log(1/\delta)$ for the smallest support size $K_{\min}\ge 1$ of
interest the call aborts, an event of probability $\le\delta$ by the same concentration. For $K=O(1)$ the
cover construction has already found all support cells explicitly (\cref{lem:packing}'s small-$b$ case),
and $K$ is known exactly with no sampling. Until \cref{sec:support-call} we treat $K$ as known exactly and
account for the $\kappa$-perturbation at the end.

\subsection{Overview}\label{sec:support-overview}

By the cluster-count integral~\eqref{eq:cluster-integral}, $\Wq=\int_0^\infty(\cQ(t)-1)\,dt$, so it
suffices to estimate the component-count curve $\cQ$ of the support and integrate. The difficulty is one
of \emph{rare, heavy} structure. Consider a dense $\sqrt K\times\sqrt K$ block of support points together
with a few far-apart satellites; the satellites are a vanishing fraction of $Q$ yet they carry a constant
fraction of $\Wq$. Any estimator that learns $\cQ$ by sampling $\poly\log K$ support \emph{points} and
exploring their neighborhoods will, with high probability, never see a satellite, and so it underestimates
$\Wq$ by a constant factor. Recovering the heavy tail forces $\Omega(\sqrt K)$ point samples, which is the
barrier we must avoid charging at every scale.

The fix is to sample \emph{space}, not mass. At a threshold $r$ we impose a grid of side $a=\Theta(\sigma
r)$ and form the \emph{scale graph} $\Gamma_r$ whose vertices are the nonempty $a$-cells and whose edges
join cells within distance $r$; its component count $\cG(r)$ brackets $\cQ$ at nearby thresholds
(\cref{sec:support-filtration}). To estimate $\cG(r)$ we sample candidate cells \emph{uniformly from a
spatial universe that includes empty cells}, and on an occupied sample we run a min-rank exploration that
returns a nonzero value only at the unique \emph{leader} of its component. A far satellite occupies its
own $a$-cell at its own scale, so it is one spatial cell among the candidate universe and is hit with the
right probability; emptiness is cheap to test, and empty samples contribute zero. This is the empty-cell
leader estimator of \cref{sec:support-estimator}; it is the one new primitive of the paper.

Two quantities then control the cost, and separating them is what makes the bound $\Otil(\sqrt K)$ and not
$\Otil(K)$. The \emph{number of trials} needed to estimate $\cG(r)$ to a target additive accuracy is
governed by the estimator's variance, which we show is at most $M c$ where $M$ is the candidate-universe
size and $c=\cG(r)$ is the \emph{component count} -- not the number of occupied cells. The \emph{per-trial
exploration cost}, on the other hand, does scale with the number of occupied cells $N$, but only through
the crude bound $N\le K$. Summed over a geometric sequence of scales, the packing inequality $\Wq\ge
b\delta/16$ of the active cover (\cref{sec:support-cover}) collapses both sums to $\Otil(\sqrt K)$.

\paragraph{A warm-up.}
The dense-block-plus-satellites instance of \cref{sec:overview} shows the two cost drivers concretely.
Take a $\sqrt K\times\sqrt K$ block of occupied unit cells together with $s$ satellites placed at pairwise
distances $\Theta(D)$ for some large $D$, so $K'=K+s$ cells in all and $\Wq=\Theta(\sqrt K)+\Theta(sD)$.
At a scale $r\ll D$ the block is one cluster and each satellite is its own, so $c=\cG(r)=s+1$. A mass
sample of $\poly\log$ points lands in a satellite with probability $\Theta(s/K')$ and misses all of them
unless $\Omega(K'/s)$ points are drawn; this is the $\sqrt n$ trap. The leader estimator instead draws
from the spatial universe of all $M=\Theta(K')$ candidate cells, so a satellite is hit with probability
$\Theta(1/M)$ like any other cell, and by \eqref{eq:ov-var} the variance is $Mc=\Theta(K' s)$, so
$\Otil(M c/\alpha^2)$ trials at additive accuracy $\alpha=\Theta(\eps\Wq/(\text{\#scales}\cdot r))$ suffice.
The exploration cost stays bounded because exploring the block component touches $O(\log K)$ cells in
expectation, not all $K$. The packing bound converts ``$M c$ trials, each $O(\log K)$ work'' into
$\Otil(\sqrt K)$ once summed over scales; \cref{sec:app-accounting} carries this out for arbitrary
supports.

\subsection{The active cover and its packing bound}\label{sec:support-cover}

The cover is a set of $\Theta(\sqrt K)$ disjoint cells, occupied by $Q$, that we can find cheaply and that
certifies a lower bound on $\Wq$. Let $q=\ceil{\sqrt K}$. Impose the quadtree on a bounding square of $Q$
and descend level by level, at each level counting how many cells are nonempty (one range count per
candidate cell; only the children of nonempty cells need be examined, so a level with $m$ nonempty cells
costs $O(m)$ counts). Stop at the first dyadic level $\mathcal B$ at which at least $q$ cells are nonempty;
let $b=\abs{\mathcal B}$ and let $\delta$ be the common side of those cells. The previous level had fewer
than $q$ nonempty cells, and quartering a cell creates at most four nonempty children, so the new count is
less than $4q$; together with the stopping rule,
\begin{equation}\label{eq:cover-size}
  q\le b<4q,\qquad\text{i.e.}\qquad b=\Theta(\sqrt K).
\end{equation}
Each level visited has $O(q)=O(\sqrt K)$ nonempty cells, and there are $O(\log\Delta)$ levels, so the
whole construction costs $\Otil(\sqrt K)$ range counts on $P$ (\cref{sec:support-oracle}). The cover
supports the packing bound that drives the whole accounting.

\begin{lemma}[Active-cover packing]\label{lem:packing}
$\Wq\ge b\delta/16$.
\end{lemma}

\begin{proof}[Proof sketch]
Pick one support point in each block of $\mathcal B$ and four-color the blocks by the parities of their
grid indices. Points chosen from distinct same-color blocks are at distance at least $\delta$, so some
color class $S$ of at least $b/4$ points has $\wt(\MST(S))\ge\delta(\abs S-1)$. For any $S\subseteq Q$,
contracting $\MST(Q)$ onto $S$ and shortcutting gives $\wt(\MST(S))\le 2\Wq$. Combining,
$\Wq\ge\delta(b-4)/8\ge b\delta/16$ for $b\ge 8$; the few small cases are handled by direct enumeration.
Full details in \cref{sec:app-lemma3}.
\end{proof}

\subsection{The empty-cell leader estimator}\label{sec:support-estimator}

Let $\mathcal U$ be a known family of $M$ candidate cells that contains every vertex of a graph $\Gamma$
on occupied cells, and suppose $\Gamma$ has $c$ components. One \emph{trial} draws $X$ uniformly from
$\mathcal U$ and outputs $Z$ as follows: if $X$ is empty, $Z=0$; otherwise assign independent uniform
ranks to occupied cells (lazily) and explore the component of $X$ in $\Gamma$, stopping as soon as a cell
of smaller rank than $X$ is met. Output $Z=M$ if no such cell exists (so $X$ is the minimum-rank cell, the
\emph{leader}, of its component), and $Z=0$ otherwise.

\begin{algorithm}[t]
\caption{One trial of the empty-cell leader estimator}\label{alg:leader}
\KwIn{candidate universe $\mathcal U$ of $M$ cells; graph $\Gamma$ on the occupied cells}
\KwOut{$Z\in\{0,M\}$ with $\E Z=c$, the component count of $\Gamma$}
draw $X$ uniformly from $\mathcal U$\;
\lIf{$X$ is empty (one range count)}{\Return $Z=0$}
assign $X$ and the cells it reaches independent uniform ranks (lazily)\;
explore the component of $X$ in $\Gamma$ in a fixed order independent of the ranks (breadth-first, neighbors ordered by cell identifier), halting once a cell of rank $<\mathrm{rank}(X)$ appears\;
\lIf{no smaller-rank cell was met ($X$ is the leader)}{\Return $Z=M$}
\lElse{\Return $Z=0$}
\end{algorithm}

\begin{lemma}[Leader estimator]\label{lem:leader}
$\E Z=c$ and $\Var(Z)=Mc-c^2\le Mc$.
\end{lemma}

\begin{proof}
For any fixed rank assignment exactly one occupied cell per component is its leader, so exactly $c$ of the
$M$ candidate cells return $Z=M$ and the rest return $0$. Hence $\Pr[Z=M]=c/M$, giving $\E Z=c$ and
$\E Z^2=M^2\cdot c/M=Mc$, so $\Var(Z)=Mc-c^2\le Mc$. The occupied-cell count never enters.
\end{proof}

Given an upper bound $U\ge c$, a median-of-means average of $k=O\!\bigl((MU/\alpha^2)\log(1/\delta)\bigr)$
trials estimates $c$ to additive error $\alpha$ with probability $1-\delta$. The per-trial exploration is
cheap: a sampled cell is occupied with probability $N/M$ (where $N$ is the number of occupied cells), and
an occupied trial in a component of size $s$ examines in expectation at most $1+H_{s-1}=O(\log K)$ cells.
The exploration order $v_1=X,v_2,\dots$ is the breadth-first order of \cref{alg:leader}, fixed before the
ranks are revealed; cell $v_j$ is examined only when $X$ has the smallest rank among $v_1,\dots,v_{j-1}$
(otherwise a smaller-rank cell appeared earlier and stopped the search), an event of probability
$1/(j-1)$ by exchangeability of the independent uniform ranks. Each examined cell costs one range count to test for
occupancy and $O(\sigma^{-2})$ counts to find its neighbors (\cref{sec:support-filtration}). The two cost
drivers -- the trial count $k$ (through the variance bound $MU$) and the expected number of explored cells
(through $N$) -- are accounted separately in \cref{sec:app-lemma3}, using $U\ge c$ for the first and
$N\le K$ for the second.

\subsection{Scale graphs and the filtration}\label{sec:support-filtration}

At a threshold $r$ we want a grid fine enough that grouping points into cells barely changes the
threshold-$r$ component structure, yet coarse enough that few cells lie within distance $r$ of any one cell
(so neighborhoods are cheap). A cell side $a=\Theta(\sigma r)$ achieves both, except when $r$ is so small
that $\sigma r$ would fall below the support resolution $h$, in which case we simply use the support grid
itself. Concretely, fix $\beta\in(0,\tfrac1{16})$ and put $\sigma=\beta/32$. For a threshold $r\ge h$ let
$a(r)$ be a dyadic multiple of $h$ with $a(r)=h$ in the \emph{floor regime} $h>\sigma r/128$, and
$\sigma r/128\le a(r)\le\sigma r/64$ in the \emph{scaling regime} otherwise. The scale graph $\Gamma_r$ has
the nonempty $a(r)$-cells as vertices, with two cells adjacent when their Euclidean set-distance is at most
$r$ (for $a(r)=h$, when their centers are within $r$). Because
$a(r)=\Theta(\sigma r)$ in the non-floor regime and $r<128h/\sigma$ in the floor regime, only
$O(\sigma^{-2})$ cells lie within distance $r$ of any fixed cell, so $\Gamma_r$ has degree
$O(\sigma^{-2})$ and each neighbor test is one range count on an aligned rectangle.

\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.62, every node/.style={font=\scriptsize}]
  \draw[step=1cm, gray!35, thin] (0,0) grid (9,5);
  % component A (left)
  \foreach \x/\y in {0/0,1/0,0/1,1/2,2/1} {\fill[blue!18] (\x,\y) rectangle (\x+1,\y+1);}
  % component B (top middle)
  \foreach \x/\y in {3/3,4/3,4/4,5/3} {\fill[green!18] (\x,\y) rectangle (\x+1,\y+1);}
  % isolated satellites
  \fill[orange!25] (7,1) rectangle (8,2);
  \fill[orange!25] (8,4) rectangle (9,5);
  % edges within distance r (schematic) inside each component
  \draw[blue!60,thick] (0.5,0.5)--(1.5,0.5) (0.5,0.5)--(0.5,1.5) (0.5,1.5)--(1.5,2.5) (1.5,2.5)--(2.5,1.5);
  \draw[green!50!black,thick] (3.5,3.5)--(4.5,3.5) (4.5,3.5)--(4.5,4.5) (4.5,3.5)--(5.5,3.5);
  \node[blue!60!black] at (1.0,-0.5) {component};
  \node[orange!75!black] at (7.5,0.5) {isolated};
  \node[orange!75!black] at (8.5,3.5) {isolated};
  \draw[<->] (6.0,0.3)--(7.0,0.3); \node at (6.5,-0.05) {$\le r$?};
\end{tikzpicture}
\caption{A scale graph $\Gamma_r$ at threshold $r$ on side-$a(r)$ cells. Vertices are nonempty cells;
two are adjacent when within Euclidean distance $r$ (degree $O(\sigma^{-2})$). Here $\cG(r)=4$: two
multi-cell components and two isolated satellite cells. The leader estimator estimates this component
count by spatial sampling, and \eqref{eq:interleave} brackets it against the support's curve $\cQ$.}
\label{fig:scalegraph}
\end{figure}

Writing $\cG(r)$ for the number of components of $\Gamma_r$ (\cref{fig:scalegraph}), the scale graph
brackets the support's component-count curve: with $r^+=(1+\sigma)r$,
\begin{equation}\label{eq:interleave}
  \cQ(r^+)\ \le\ \cG(r)\ \le\ \cQ(r).
\end{equation}
The upper bound holds because every Euclidean $r$-edge of $Q$ induces an edge of $\Gamma_r$ (then
intra-cell contraction only merges components); the lower bound because cells adjacent in $\Gamma_r$ hold
points within $r+2\sqrt2\,a(r)<r^+$. Combined with~\eqref{eq:components-edges}, $\cG(r)\le 1+\Wq/r$, so
under any guess $G\ge\Wq$ the deterministic bound $U(r)=1+G/r$ satisfies $U(r)\ge\cG(r)$ -- a bound on the
\emph{component} count, which is what \cref{lem:leader} needs.

\subsection{The candidate universe}\label{sec:support-cand}

The leader estimator needs a universe $\mathcal U$ of candidate cells that we can both sample uniformly
and recognize, and that contains every occupied cell of the current scale graph. At a fine scale
($a=a(r)\le\delta$) every nonempty $a$-cell of $Q$ lies inside a block of the active cover $\mathcal B$:
indeed any support point sits in some $\delta$-block, and its $a$-cell is a subcell of that block. So we
take $\mathcal U$ to be the set of all $a$-subcells of the $b$ active blocks, of size
\begin{equation}\label{eq:Msize}
  M(r)=b\,(\delta/a)^2\le C_M\,\frac{b\delta^2}{\sigma^2 r^2},
\end{equation}
using $a=\Theta(\sigma r)$. A uniform candidate is generated with no queries at all -- pick one of the $b$
active blocks uniformly, then one of its $(\delta/a)^2$ subcells uniformly -- and a single range count
then tests whether that $a$-cell is occupied. This is exactly the spatial sampling the estimator wants:
the universe is known offline, empties are cheap, and an occupied cell is found in proportion to its share
of $\mathcal U$, independent of how many points it holds. At a coarse scale ($a>\delta$) there is nothing
to sample: the occupied $a$-cells are precisely the $\le b$ distinct $a$-ancestors of the active blocks,
so $\Gamma_r$ and its component count $\cG(r)$ are read off the cover directly, with no further queries and
no recursion into the points. Computing $\cG(r)$ takes $O(b\sigma^{-2})$ time, not $O(b^2)$: hash the
occupied $a$-cells by coordinates and, for each, probe its $O(\sigma^{-2})$ neighbor positions in the
hash table to build the adjacency of $\Gamma_r$, then count components by union-find.

\subsection{The additive call and removing the guess}\label{sec:support-call}

Assume a guess $G\ge\Wq$. Set $r_0=\max\set{h,\beta G/(32K)}$, $r_j=r_0(1+\sigma)^j$, and $d_j=\sigma
r_j$, stopping at $r_T>\sqrt2\Delta\ge\diam(Q)$, so $T=O(\sigma^{-1}\log(\Delta/h))$. At each fine scale ($a(r_j)\le
\delta$) estimate $c_j\eqdef\cG(r_j)$ to additive accuracy $\alpha_j=\beta G/(32\,T\,\sigma r_j)$ with the
leader estimator over the candidate universe of \cref{sec:support-cand}, using $U_j=1+G/r_j$; coarse
scales are computed exactly from the cover with no further queries. Output
\begin{equation}\label{eq:additive-call}
  \widehat W(G)=(K-1)h+\sum_{j=0}^{T-1} d_j\,(\widehat c_j-1).
\end{equation}

\begin{lemma}[Additive call]\label{lem:additive}
If $G\ge\Wq$, then with probability $\ge 5/6$ the estimator~\eqref{eq:additive-call} satisfies
$\abs{\widehat W(G)-\Wq}\le\beta G$ and uses $\Otil_\beta(\sqrt K)$ range-counting queries.
\end{lemma}

The estimator \eqref{eq:additive-call} is a Riemann sum for the cluster-count integral $\Wq=\int_0^\infty
(\cQ-1)$, and its error decomposes into three pieces, each kept to $\beta G/32$ by the choices above. The
$(K-1)h$ term is the exact contribution of $[0,h]$ (distinct centers are $\ge h$ apart, so $\cQ=K$ there);
the \emph{omitted interval} $[h,r_0]$ contributes at most $Kr_0=\beta G/32$ by the choice of $r_0$; the
\emph{Riemann gap} between the discretized sum and the integral is at most the total staircase drop
$\sum_j d_j(\cQ(r_j)-\cQ(r_{j+1}))\le\sigma\Wq\le\beta G/32$, since an MST edge of length $\ell\in(r_j,
r_{j+1}]$ contributes $d_j=\sigma r_j\le\sigma\ell$; and the \emph{statistical error} $\sum_j d_j\abs{\widehat
c_j-c_j}\le\sum_j d_j\alpha_j$ telescopes to $\beta G/32$ because $d_j\alpha_j=\sigma r_j\cdot\beta G/(32T
\sigma r_j)=\beta G/(32T)$ is the same at every scale. Summing the three gives $\abs{\widehat W(G)-\Wq}\le
3\beta G/32<\beta G$. The cost is the two-part sum of \cref{sec:support-overview}, both branches equal to
$\Otil_\beta(\sqrt K)$ via \cref{lem:packing}. Both the accuracy and the cost are carried out in full in
\cref{sec:app-lemma3}.

To remove the guess, start from the valid bound $U_Q=\sqrt2(K-1)\Delta\ge(K-1)\diam(Q)\ge\Wq$ (computable
without knowing $\diam(Q)$, since $\diam(Q)\le\sqrt2\Delta$) and run \cref{lem:additive} with
$\beta_0=\tfrac1{32}$ on $G=U_Q,U_Q/2,U_Q/4,\dots$, stopping at the first $G$ with $\widehat W(G)\ge G/4$.
A short calculation (\cref{sec:app-lemma3}) shows the selected $G_0$ satisfies $\Wq\le G_0<8\Wq$; a final
call with $\beta=\eta/16$ then yields $\abs{\widehat\Wq-\Wq}\le\tfrac\eta2\Wq$, and amplification over the
$O(\log(\Delta/h))$ guesses keeps the total cost $\Otil_\eta(\sqrt K)$. Replacing $K$ by an estimate of
relative error $O(\eta)$ (\cref{sec:support-oracle}) perturbs every bound by a $1+O(\eta)$ factor and adds
the $n/K$ term, completing \cref{lem:support}.
